---
title: "SNV Heatmap"
author: "Stacy Hung"
date: "March 26, 2019"
output: html_document
---

This script is used to generate plots for visualizing mutation data for the GZL exomes dataset.

## Load packages

```{r}
# NB: if you need to install this package, install it using devtools from the github source!!!
# The bioconductor version of GenVisR is buggy (not compatible with required libraries)
# Installing GenVisR may also require the installation of the VariantAnnotation pacakge (from bioconductor)
source("https://bioconductor.org/biocLite.R")
biocLite("VariantAnnotation")
library("VariantAnnotation")

install.packages("devtools")
library(devtools)
install_github("griffithlab/GenVisR")
library("GenVisR")

source("https://bioconductor.org/biocLite.R")
biocLite("maftools")
library(maftools)

library(gridExtra)
library(dplyr)
library(plyr)
```

## Shape data to get frequently mutated (e.g. 10%+) based on pathological group
```{r}
library(dplyr)

cases.mutated.pathgp <- read.table("/Volumes/shung/projects/gzl_exomes/data/Path_group_visualization/patient_gene_path_group.txt", sep = "\t", header = TRUE)

# get numbers of patients in each group
patients.group0 <- subset(cases.mutated.pathgp, cases.mutated.pathgp$Path_group == "0 = cHL")
NUM_GROUP0 <- length(unique(patients.group0$Tumor_Sample_Barcode))
patients.group1_2 <- subset(cases.mutated.pathgp, cases.mutated.pathgp$Path_group == "1 = cHL-interm" | cases.mutated.pathgp$Path_group == "2 = LBCL-interm")
NUM_GROUP1_2 <- length(unique(patients.group1_2$Tumor_Sample_Barcode))
patients.group3 <- subset(cases.mutated.pathgp, cases.mutated.pathgp$Path_group == "3 = LBCL")
NUM_GROUP3 <- length(unique(patients.group3$Tumor_Sample_Barcode))

counts.mutated.pathgp <- as.data.frame(table(cases.mutated.pathgp$Path_group, cases.mutated.pathgp$Hugo_Symbol))
colnames(counts.mutated.pathgp) <- c("Path_group", "gene", "num_mutated")

# extract genes that are mutated and specific to pathological groups that need to be filtered
counts.mutated.pathgp.0 <- subset(counts.mutated.pathgp, 
                                  counts.mutated.pathgp$Path_group == "0 = cHL" &
                                    counts.mutated.pathgp$num_mutated > 0)
counts.mutated.pathgp.1_2 <- subset(counts.mutated.pathgp, 
                                    (counts.mutated.pathgp$Path_group == "1 = cHL-interm" | 
                                       counts.mutated.pathgp$Path_group == "2 = LBCL-interm") &
                                      counts.mutated.pathgp$num_mutated > 0)
counts.mutated.pathgp.3 <- subset(counts.mutated.pathgp,
                                  counts.mutated.pathgp$Path_group == "3 = LBCL" &
                                   counts.mutated.pathgp$num_mutated > 0)

# calculate proportion of mutated cases
counts.mutated.pathgp.0$proportion <- counts.mutated.pathgp.0$num_mutated / NUM_GROUP0*100
counts.mutated.pathgp.1_2$proportion <- counts.mutated.pathgp.1_2$num_mutated / NUM_GROUP1_2*100

PERCENT_MUTATED = 10

# now get subset of genes that are mutated in 10% or higher for each group
counts.mutated.pathgp.0_cutoff <- subset(counts.mutated.pathgp.0, counts.mutated.pathgp.0$proportion >= PERCENT_MUTATED)
counts.mutated.pathgp.1_2_cutoff <- subset(counts.mutated.pathgp.1_2, counts.mutated.pathgp.1_2$proportion >= PERCENT_MUTATED)
pathgp_genes.10_plus <- unique(c(as.vector(counts.mutated.pathgp.0_cutoff$gene),as.vector(counts.mutated.pathgp.1_2_cutoff$gene),
                                 as.vector(counts.mutated.pathgp.3$gene)))

write.table(pathgp_genes.10_plus, "/Volumes/shung/projects/gzl_exomes/data/Path_group_visualization/pathgp_genes.10_plus.txt", quote = FALSE, row.names = FALSE)

# create mutation matrix for filtered gene list
mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.oncoplot.maf", sep = "\t", header = TRUE, fill = TRUE)
mutations.df <- filter(mutations.df, mutations.df$Hugo_Symbol %in% pathgp_genes.10_plus)
mutations.df$Hugo_Symbol <- factor(mutations.df$Hugo_Symbol)
mutations.df <- unique(mutations.df[c("Hugo_Symbol", "Tumor_Sample_Barcode")])
mutation.matrix <- table(mutations.df$Tumor_Sample_Barcode, mutations.df$Hugo_Symbol)
write.table(mutation.matrix, "/Volumes/shung/projects/gzl_exomes/data/Path_group_visualization/mutation_matrix-path_groups_10+_freq_mutated.txt", sep = "\t", quote = FALSE)
```


## Shape data to get frequently (e.g. 10%+) mutated EBV+ and EBV- cases.

```{r}
cases.mutated.ebv_status <- read.table("/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/EBV_status_mutated_cases.oncoplot.txt", sep = "\t", header = TRUE)

counts.mutated.ebv_status <- as.data.frame(table(cases.mutated.ebv_status$EBER, cases.mutated.ebv_status$Hugo_Symbol))
colnames(counts.mutated.ebv_status) <- c("EBV_status", "gene", "num_mutated")
counts.mutated.ebv_status.POS <- subset(counts.mutated.ebv_status, counts.mutated.ebv_status$EBV_status == "Positive" & counts.mutated.ebv_status$num_mutated > 0)
counts.mutated.ebv_status.NEG <- subset(counts.mutated.ebv_status, counts.mutated.ebv_status$EBV_status == "Negative" & counts.mutated.ebv_status$num_mutated > 0)

# calculate proportion (% cases mutated) (22 cases are EBV-, 7 cases are EBV+)
counts.mutated.ebv_status.POS$proportion <- counts.mutated.ebv_status.POS$num_mutated / 7 * 100
counts.mutated.ebv_status.NEG$proportion <- counts.mutated.ebv_status.NEG$num_mutated / 22 * 100

# get subset of genes that are mutated in a given % of each of the EBV groups
PERCENT_MUTATED = 10
#PERCENT_MUTATED = 20
counts.mutated.ebv_status.NEG.cutoff <- subset(counts.mutated.ebv_status.NEG, counts.mutated.ebv_status.NEG$proportion >= PERCENT_MUTATED)
counts.mutated.ebv_status.POS.cutoff <- subset(counts.mutated.ebv_status.POS, counts.mutated.ebv_status.POS$proportion >= PERCENT_MUTATED)
EBV_genes.10_plus <- unique(c(as.vector(counts.mutated.ebv_status.NEG.cutoff$gene), 
                              as.vector(counts.mutated.ebv_status.POS.cutoff$gene)))
#EBV_genes.20_plus <- unique(c(as.vector(counts.mutated.ebv_status.NEG.cutoff$gene),       as.vector(counts.mutated.ebv_status.POS.cutoff$gene)))

# create EBV-filtered dataset (mutated in 10%+ for each subgroup)
mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.oncoplot.maf", sep = "\t", header = TRUE, fill = TRUE)
mutations.df <- filter(mutations.df, mutations.df$Hugo_Symbol %in% EBV_genes.10_plus)
mutations.df$Hugo_Symbol <- factor(mutations.df$Hugo_Symbol)
mutations.df <- unique(mutations.df[c("Hugo_Symbol", "Tumor_Sample_Barcode")])
mutation.matrix <- table(mutations.df$Tumor_Sample_Barcode, mutations.df$Hugo_Symbol)
#write.table(mutation.matrix, "/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/mutation_matrix-EBV_10+_freq_mutated.txt", sep = "\t", quote = FALSE)

```

## One time: Process maf file so that sample ids are consistent and genes to display are those of interest (in target panel)

```{r}
# read in mutations
mutations <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.maf", sep = "\t", header = TRUE, fill = TRUE)
# filter for genes of interest
mutations <- filter(mutations, mutations$Hugo_Symbol %in% genes.target$gene)
mutations$Hugo_Symbol <- factor(mutations$Hugo_Symbol)

# get list of genes that are mutated in 2 or more cases
mutations.temp <- unique(mutations[, c("Hugo_Symbol", "Tumor_Sample_Barcode")])
cases.mutated <- as.data.frame(table(mutations.temp$Hugo_Symbol))
colnames(cases.mutated) <- c("gene", "cases_mutated")
cases.mutated <- filter(cases.mutated, cases.mutated$cases_mutated > 1)
mutations <- filter(mutations, mutations$Hugo_Symbol %in% cases.mutated$gene)

# change mapping of ids
mutations$Tumor_Sample_Barcode <- revalue(mutations$Tumor_Sample_Barcode, 
                                          c("GZ_BCC_54_T_LMD"="GZ-BCC-054",
                                            "GZ_BCC_08_T_LMD"="GZ-BCC-008",
                                            "GZ_BCC_13_T_LMD"="GZ-BCC-013",
                                            "GZ_BCC_20_T_LMD"="GZ-BCC-020",
                                            "GZ32TLMD"="GZ-032",
                                            "GZ044T-merged"="GZ-044",
                                            "GZ046T-merged"="GZ-046",
                                            "GZ048T-merged"="GZ-048",
                                            "GZ062T-merged"="GZ-062",
                                            "GZ064T"="GZ-064",
                                            "GZ068T-merged"="GZ-068",
                                            "GZ86TLMD"="GZ-086",
                                            "GZ092_TLMD_2"="GZ-092",
                                            "GZ095_TLMD_2"="GZ-095",
                                            "99T"="GZ-099",
                                            "GZ116T"="GZ-116",
                                            "GZ149TLMD"="GZ-149", 
                                            "GZ152T-merged"="GZ-152",
                                            "176T"="GZ-176",
                                            "GZ178T"="GZ-178",
                                            "GZ180_FFPE"="GZ-180",
                                            "GZ184TLMD"="GZ-184",
                                            "GZ197_TLMD"="GZ-197",
                                            "GZ230T"="GZ-230",
                                            "GZ235T"="GZ-235",
                                            "GZ267T-merged"="GZ-267",
                                            "GZ294T"="GZ-294",
                                            "GZ301T"="GZ-301",
                                            "GZ310T"="GZ-310"
                                            ))

#write.table(mutations, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.oncoplot.maf", sep = "\t", row.names = FALSE, quote = FALSE)

```

## Oncoplots

```{r}
# load genes to display for oncoplot
genes.target <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/WES_data/target_genes.FINAL_217.txt", header = FALSE)
colnames(genes.target) <- c("gene")
genes.target <- as.vector(genes.target$gene)

# load metadata
clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.mapped.txt", sep = "\t", header = TRUE)

mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.oncoplot.genes.ordered.maf", clinicalData = clinical.data)

# colours for annotation tracks
annot.colors <- list(Gender = c("Female" = "palevioletred1", "Male" = "royalblue1"),
                     EBER = c("Negative" = "black", "Positive" = "red"),
                     Mediastinal = c("Not_involved" = "grey", "Involved" = "mediumpurple"),
                     Age_yrs = c("<_45" = "bisque", ">=_45" = "bisque4"),
                     Path_group = c("0_=_cHL" = "lightsalmon",
                                   "1_=_cHL-interm" = "khaki1",
                                   "2_=_LBCL-interm" = "lightgreen", 
                                   "3_=_LBCL" = "steelblue1")
               )

# Standard mutation waterfall plot (oncoplot style), and with clinical annotations
oncoplot(maf = mutations.maf, genes = genes.target$gene, 
         clinicalFeatures = c("Path_group", "Mediastinal", "EBER", "Gender", "Age_yrs"),
         drawColBar = FALSE, 
         legendFontSize = 12, 
         showTumorSampleBarcodes = TRUE,
         annotationColor = annot.colors)

# order by path group (in specific order)
oncoplot(maf = mutations.maf, genes = genes.target$gene, 
         clinicalFeatures = c("Path_group", "Mediastinal", "EBER", "Gender", "Age_yrs"),
         drawColBar = FALSE, 
         legendFontSize = 12, 
         showTumorSampleBarcodes = TRUE,
         annotationColor = annot.colors, sortByAnnotation = TRUE, 
         annotationOrder = c("0_=_cHL", "1_=_cHL-interm", "2_=_LBCL-interm", "3_=_LBCL"))

# order by EBV-status
oncoplot(maf = mutations.maf, genes = genes.target$gene, 
         clinicalFeatures = c("EBER", "Path_group", "Mediastinal", "Gender", "Age_yrs"),
         drawColBar = FALSE, 
         legendFontSize = 12, 
         showTumorSampleBarcodes = TRUE,
         annotationColor = annot.colors, sortByAnnotation = TRUE)

# show only genes mutated in EBV+ patients
oncoplot(maf = mutations.maf, genes = c("ACTB", "ATP13A4", "DHX58", "STAT3"), 
         clinicalFeatures = c("EBER", "Path_group", "Mediastinal", "Gender", "Age_yrs"),
         drawColBar = FALSE, 
         legendFontSize = 12, 
         showTumorSampleBarcodes = TRUE,
         annotationColor = annot.colors, sortByAnnotation = TRUE, removeNonMutated = FALSE)

# show genes mutated in 10%+ of each EBV category
oncoplot(maf = mutations.maf, genes = EBV_genes.10_plus, 
         clinicalFeatures = c("EBER", "Path_group", "Mediastinal", "Gender", "Age_yrs"),
         drawColBar = FALSE, 
         legendFontSize = 12, 
         showTumorSampleBarcodes = TRUE,
         annotationColor = annot.colors, sortByAnnotation = TRUE)

# show genes mutated in 20%+ of each EBV category
oncoplot(maf = mutations.maf, genes = EBV_genes.20_plus, 
         clinicalFeatures = c("EBER", "Path_group", "Mediastinal", "Gender", "Age_yrs"),
         drawColBar = FALSE, 
         legendFontSize = 12, 
         showTumorSampleBarcodes = TRUE,
         annotationColor = annot.colors, sortByAnnotation = TRUE, removeNonMutated = FALSE)

```

## EBV-specific oncoplots

```{r}
# to get latest version of maftools, install from github
library("devtools")
install_github(repo = "PoisonAlien/maftools")
library(maftools)

clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.mapped.txt", sep = "\t", header = TRUE)

mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.oncoplot.genes.ordered.maf", clinicalData = clinical.data)

# plot by EBV+ and EBV- sorted cohorts
# list of genes that are mutated in at least 10% of EBV+ and EBV- cases, sorted by gene
genes.ordered <- read.table("/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/oncoplot-genes-ordered_by_ebv_ebv+_ebv_neg.txt", header = TRUE)
genes.ordered <- as.vector(genes.ordered$gene)

# colours for annotation tracks
annot.colors <- list(Gender = c("Female" = "palevioletred1", "Male" = "royalblue1"),
                     EBER = c("Negative" = "black", "Positive" = "red"),
                     Mediastinal = c("Not_involved" = "grey", "Involved" = "mediumpurple"),
                     Age_yrs = c("<_45" = "bisque", ">=_45" = "bisque4"),
                     Path_group = c("0_=_cHL" = "lightsalmon",
                                   "1_=_cHL-interm" = "khaki1",
                                   "2_=_LBCL-interm" = "lightgreen", 
                                   "3_=_LBCL" = "steelblue1")
               )

patients.ordered <- c("GZ-062", "GZ-267", "GZ-068", "GZ-086", "GZ-230", "GZ-092", "GZ-BCC-054", "GZ-116", "GZ-099", "GZ-BCC-020", "GZ-310", "GZ-149", "GZ-044", "GZ-184", "GZ-046", "GZ-178", "GZ-BCC-013", "GZ-176", "GZ-095", "GZ-152", "GZ-048", "GZ-180", "GZ-BCC-008", "GZ-301", "GZ-032", "GZ-294", "GZ-064", "GZ-197", "GZ-235")

oncoplot(maf = mutations.maf, 
         sampleOrder = patients.ordered,
         genes = genes.ordered, 
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 1, annotationFontSize = 1,
         clinicalFeatures = c("EBER", "Path_group", "Mediastinal", "Gender", "Age_yrs"),
         annotationColor = annot.colors, 
         showTumorSampleBarcodes = TRUE, 
         drawColBar = FALSE)

  #legendFontSize = 1, 

```

## Path group-specific oncoplots

```{r}
# to get latest version of maftools, install from github
library("devtools")
install_github(repo = "PoisonAlien/maftools")
library(maftools)

clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.mapped.txt", sep = "\t", header = TRUE)

mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.oncoplot.maf", clinicalData = clinical.data)

# plot by EBV+ and EBV- sorted cohorts
# list of genes that are mutated in at least 10% of EBV+ and EBV- cases, sorted by gene
genes.ordered <- read.table("/Volumes/shung/projects/gzl_exomes/data/Path_group_visualization/genes.ordered.txt", header = TRUE)
genes.ordered <- as.vector(genes.ordered$gene)

patients.ordered <- read.table("/Volumes/shung/projects/gzl_exomes/data/Path_group_visualization/patients.ordered.txt", header = TRUE)
patients.ordered <- as.vector(patients.ordered$patient)

# colours for annotation tracks
annot.colors <- list(Gender = c("Female" = "palevioletred1", "Male" = "royalblue1"),
                     EBER = c("Negative" = "black", "Positive" = "red"),
                     Mediastinal = c("Not_involved" = "grey", "Involved" = "mediumpurple"),
                     Age_yrs = c("<_45" = "bisque", ">=_45" = "bisque4"),
                     Path_group = c("0_=_cHL" = "lightsalmon",
                                   "1_=_cHL-interm" = "khaki1",
                                   "2_=_LBCL-interm" = "lightgreen", 
                                   "3_=_LBCL" = "steelblue1")
               )

oncoplot(maf = mutations.maf, 
         sampleOrder = patients.ordered,
         genes = genes.ordered, 
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 1, annotationFontSize = 1,
         clinicalFeatures = c("Path_group", "EBER", "Mediastinal", "Gender", "Age_yrs"),
         annotationColor = annot.colors, 
         showTumorSampleBarcodes = TRUE, 
         drawColBar = FALSE, removeNonMutated = FALSE)

  #legendFontSize = 1, 


```

## Customized oncoprint so that we can specify the order of BOTH genes and patients

```{r}
library(ComplexHeatmap)

mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.oncoplot.genes.ordered.maf", sep = "\t", header = TRUE, fill = TRUE)
```

## Somatic interactions

```{r}

# plot somatic interactions
somaticInteractions(maf = mutations.maf, genes=sig.genes, top = 50, pvalue = c(0.05, 0.01))

# oncoplots for EACH mutually exclusive / co-occurring interaction (most significant):

# mutually exclusive interactions: (30x3)
oncoplot(maf = mutations.maf, genes = c("GNA13", "IL4R"), removeNonMutated = FALSE,
                     drawRowBar = FALSE, drawColBar = FALSE, titleFontSize = 0, fontSize = 15)
oncoplot(maf = mutations.maf, genes = c("CISH", "STAT6"), removeNonMutated = FALSE,
                     drawRowBar = FALSE, drawColBar = FALSE, titleFontSize = 0, fontSize = 15)
oncoplot(maf = mutations.maf, genes = c("CXCR5", "SOCS1"), removeNonMutated = FALSE,
                     drawRowBar = FALSE, drawColBar = FALSE, titleFontSize = 0, fontSize = 15)

# co-occurring interactions:
oncoplot(maf = mutations.maf, genes = c("IRF2BP2", "ACTB"), 
         removeNonMutated = FALSE, 
         drawRowBar = FALSE, 
         drawColBar = FALSE, titleFontSize = 0, fontSize = 15)

laml.titv = titv(maf = mutations.maf, plot = FALSE, useSyn = FALSE)
# plot titv summary
plotTiTv(res = laml.titv)

```

## Plotting VAF

```{r}
library(dplyr)
library(maftools)

# load top recurrently mutated genes
supptable2 <- read.table("/Volumes/shung/projects/PMBCL_exomes/manuscript/current/supplementary/supp_table_2.txt", sep = "\t", header = TRUE, fill = TRUE)

mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired.maf")

sig.genes <- c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "GCSAM", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "IFNL3", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "IGLJ2", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT")
recurrent.genes <- as.vector(unique(supptable2$gene))

sfig4.genes <- as.vector(unique(c(sig.genes, recurrent.genes)))

# Plotting VAF for MutSigCV 50 genes
vafPlot <- plotVaf(maf = mutations.maf, genes = sig.genes, vafCol = 'VAF', 
                   flip = FALSE, top = 50, showN = FALSE)

# plot VAF for top recurrently mutated genes (n=95)
vafPlot <- plotVaf(maf = mutations.maf, genes = supptable2$gene, vafCol = 'VAF', showN = FALSE)

# Plotting VAF for the union of MutSig and recurrently mutated genes (n=120)
vafPlot <- plotVaf(maf = mutations.maf, genes = sfig4.genes, vafCol = 'VAF', showN = FALSE, top=123)

# plot VAF for recurrently mutated genes (NOT in MutSig)
supptable2.not_MutSig <- subset(supptable2, !(supptable2$mutsig == 'y'))
vafPlot <- plotVaf(maf = mutations.maf, genes = supptable2.not_MutSig$gene, vafCol = 'VAF', showN = FALSE)

```

## Lollipop plots

```{r}
library(maftools)
library(gridExtra)

# significant genes only
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_AA_Change.maf")

# ALL mutations *including* UTR mutations (and all genes)
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired_with_UTRs.maf", vc_nonSyn = c("Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site", "Translation_Start_Site","Nonsense_Mutation", "Nonstop_Mutation", "In_Frame_Del","In_Frame_Ins", "Missense_Mutation", "3'UTR", "5'UTR"))

mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered+IRF1_AA_Change.maf")

sig.genes <- c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT", "ALX1")

oncoclust.genes <- c("IRF4", "STAT6", "EZH2", "HIST1H3D", "ACTB", "BCL7A")

# create grid of lollipop plots for all significant genes

# for each significant gene, create plot
lolli.plots <- vector('list', length(sig.genes))
for (i in seq_along(sig.genes)) {
  lolli.plots[[i]] <- lollipopPlot(maf = mutations.maf, 
                          gene = sig.genes[i], 
                          AACol = 'Protein_Change', 
                          showMutationRate = TRUE, 
                          domainLabelSize = 2, 
                          defaultYaxis = FALSE
                          )
}
do.call("grid.arrange", c(lolli.plots, ncol=2))

# Structure for proteins: GCSAM, IFNL3, IGLJ2 not found

# Lollipop plots for amino acid changes - issues with retrieving protein position
GNA13.lpop <- lollipopPlot(maf = mutations.maf, 
                          gene = 'GNA13', 
                          AACol = 'Protein_Change', 
                          showMutationRate = TRUE, 
                          domainLabelSize = 3, 
                          defaultYaxis = FALSE, labelPos = c(27, 67, 68)
                          )

# lollipop plots for genes of interest: TBL1XR1, genes in the IRF pathway

```

## Oncodrive to detect cancer driver genes based on positional clustering

Oncodrive is based on the algorithm oncodriveCLUST (originally implemented in python).  The main idea is based on the fact that most of the variants in cancer-causing genes are enriched at few specific loci (aka hot-spots).  This method takes advantage of such positions to identify cancer genes.

```{r}
library(maftools)

# all mutated genes
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes.AA_change.maf")

# significant genes only
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_AA_Change.maf")

laml.sig = oncodrive(maf = mutations.maf, AACol = 'Protein_Change', minMut = 5, pvalMethod = 'zscore')

# plot the results using plotOncodrive
plotOncodrive(res = laml.sig, fdrCutOff = 0.1, useFraction = TRUE, labelSize = 6)

# the size of the points is proportional to the number of clusters found in the gene
# x-axis: number of mutations (or fraction of mutations) observed in these clusters

```


## Rainfall plot to detect hypermutation

```{r}
library(maftools)
library(gridExtra)

mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/singles_and_pairs/singles_and_pairs.snvs_and_indels.somatic_modified_effects.removed_non_genes.singles_AF_filtered.maf",
                          gisticAllLesionsFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.all_lesions.conf_90.txt", 
                          gisticAmpGenesFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.amp_genes.conf_90.txt", 
                          gisticDelGenesFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.del_genes.conf_90.txt",
                          gisticScoresFile = "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.scores.gistic")

# read in samples that are used as input into gistic (below noisiness threshold of 949)
samples <- read.table("/Volumes/shung/projects/PMBCL_exomes/gistic/input/samples-pass_949.txt")

# rainfall plot for every sample:

#os.plots <- vector('list', length(genes))
#for (i in seq_along(genes)) {
#  os.plots[[i]] <- mafSurvival(maf = mutations.maf, genes = genes[i], clinicalData = clinical.data, addInfo = TRUE, 
#             time = 'Overall.survival..y.', Status = 'CODE_OS', 
#             groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
#}
#do.call("grid.arrange", c(os.plots, ncol=2))

results <- rainfallPlot(maf = mutations.maf, tsb = c("PA019", "PA043"),
                        detectChangePoints = TRUE, 
                        fontSize = 12, 
                        pointSize = 0.6)

# Filter column in the above results indicate whether the identified segment passes the definition of “Kataegis” which are defined as those genomic segments containing six or more consecutive mutations with an average inter-mutation distance of less than or equal to 1,00 bp 5.

```

## Survival analysis

```{r}
library(maftools)
library(gridExtra)

# mutation data for all genes
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/singles_and_pairs/singles_and_pairs.snvs_and_indels.somatic_modified_effects.removed_non_genes.singles_AF_filtered.maf")

# mutation data filtered for significant genes and rituximab-treated patients only (!)
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/Mutsig/mutsig_input/mutation_file/maftools/paired_and_unpaired.snvs_and_indels.sig_genes_v3.hugo_gene.AF_filtered.gene_filtered.with_VAF.RITU_treated.maf")

clinical.data <- read.table("/Volumes/shung/projects/PMBCL_exomes/data/from_DaveTwa/meta_data_DTwa_updated.txt", sep = "\t", header = TRUE)

# TTP: CODE_TTP and "Time to progression (y)"
# OS: CODE_OS and "Overall survival (y)"
# PFS: CODE_PFS and "Progression free survival (y)"

genes = c("SOCS1", "ITPKB", "STAT6", "GNA13", "NFKBIE", "PTPN1", "ZNF217", "IL4R", "ACTB", "HIST1H3D", "CD58", "IRF8", "SGPP1", "TNFAIP3", "GCSAM", "JUNB", "ZFP36L1", "RGS1", "HIST1H2BC", "HIST1H1C", "JAK1", "MS4A1", "SRSF1", "HIST1H1B", "EZH2", "IFNL3", "GDF11", "KLF9", "IL1RAPL2", "SERPINB3", "VMP1", "MYC", "LHFPL3", "TSPAN19", "IRF2BP2", "DDX3X", "CIITA", "IRF4", "CXCR5", "EPHA7", "IL13RA1", "CISH", "SPINK1", "TRAF3", "HIST1H4E", "HIST1H4J", "MAGEC2", "IRX2", "FPGT")
# omitting IGLJ2A
jak.genes = c("CISH", "IL4R", "IL13RA1", "STAT6", "SOCS1", "PTPN1")

#i<-1
#genes[1] <- irf.genes

# create grid of plots for OS

# for each significant gene, create survival plot
os.plots <- vector('list', length(genes))
for (i in seq_along(genes)) {
  os.plots[[i]] <- mafSurvival(maf = mutations.maf, 
                               genes = genes[i], 
                               clinicalData = clinical.data, 
                               addInfo = TRUE, 
                               time = 'Overall.survival..y.', 
                               Status = 'CODE_OS', 
                               groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
}
do.call("grid.arrange", c(os.plots, ncol=2))

# create grid of plots for TTP

# for each significant gene, create survival plot
ttp.plots <- vector('list', length(genes))
for (i in seq_along(genes)) {
  ttp.plots[[i]] <- mafSurvival(maf = mutations.maf, 
                                genes = genes[i], 
                                clinicalData = clinical.data, 
                                addInfo = TRUE, 
                                time = 'Time.to.progression..y.', 
                                Status = 'CODE_TTP', 
                                groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
}
do.call("grid.arrange", c(ttp.plots, ncol=2))

# create grid of plots for PFS

# for each significant gene, create survival plot
pfs.plots <- vector('list', length(genes))
for (i in seq_along(genes)) {
  pfs.plots[[i]] <- mafSurvival(maf = mutations.maf, genes = genes[i], clinicalData = clinical.data, addInfo = TRUE, 
             time = 'Progression.free.survival..y.', Status = 'CODE_PFS', 
             groupNames = c(paste(genes[i],"-Mutated", sep = ""), "WT"))
}
do.call("grid.arrange", c(pfs.plots, ncol=2))


# create separate plots where genes exhibit significant differences in survival between MUT and WT
mafSurvival(maf = mutations.maf, genes = "NFKBIE", clinicalData = clinical.data, 
                                   addInfo = TRUE, time = 'Overall.survival..y.', Status = 'CODE_OS', 
                                   groupNames = c("OS: NFKBIE-Mutated", "WT"), showConfInt = FALSE)
mafSurvival(maf = mutations.maf, genes = "ITPKB", clinicalData = clinical.data, 
                                   addInfo = TRUE, time = 'Time.to.progression..y.', Status = 'CODE_TTP', 
                                   groupNames = c("TTP: ITPKB-Mutated", "WT"), showConfInt = FALSE)
mafSurvival(maf = mutations.maf, genes = "ITPKB", clinicalData = clinical.data, 
                                   addInfo = TRUE, time = 'Progression.free.survival..y.', Status = 'CODE_PFS', 
                                   groupNames = c("PFS: ITPKB-Mutated", "WT"), showConfInt = FALSE)


#hist1h3d.plots <- vector('list', 3)
#hist1h3d.plots[[1]] <- mafSurvival(maf = mutations.maf, genes = "HIST1H3D", clinicalData = clinical.data, 
#                                   addInfo = TRUE, time = 'Overall.survival..y.', Status = 'CODE_OS', 
#                                   groupNames = c("OS: HIST1H3D-Mutated", "WT"), showConfInt = FALSE)
#hist1h3d.plots[[2]] <- mafSurvival(maf = mutations.maf, genes = "HIST1H3D", clinicalData = clinical.data, 
#                                   addInfo = TRUE, time = 'Time.to.progression..y.', Status = 'CODE_TTP', 
#                                   groupNames = c("TTP: HIST1H3D-Mutated", "WT"), showConfInt = FALSE)
#hist1h3d.plots[[3]] <- mafSurvival(maf = mutations.maf, genes = "HIST1H3D", clinicalData = clinical.data, 
#                                   addInfo = TRUE, time = 'Progression.free.survival..y.', Status = 'CODE_PFS',
#                                   groupNames = c("PFS: HIST1H3D-Mutated", "WT"), showConfInt = FALSE)
#do.call("grid.arrange", c(hist1h3d.plots, ncol=3))


```

## GISTIC visualization

```{r}
library(devtools)
install_github("PoisonAlien/maftools")
library(maftools)

all.lesions <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.all_lesions.conf_90.txt"
amp.genes <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.amp_genes.conf_90.txt"
del.genes <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.del_genes.conf_90.txt"
scores.gistic <- "/Volumes/shung/projects/PMBCL_exomes/gistic/output/drop_low_coverage-filtered_949-arm_peel-threshold_0.3-fdr_0.05-max_seg_1000-FINAL_FINAL_FINAL/gistic_segments.scores.gistic"

#source("/Volumes/shung/projects/PMBCL_exomes/gistic/scripts/gisticChromPlot.R")
#source("/Volumes/shung/projects/PMBCL_exomes/gistic/scripts/readSegs.R")

laml.gistic <- maftools::readGistic(gisticAllLesionsFile = all.lesions, 
                          gisticAmpGenesFile = amp.genes, 
                          gisticDelGenesFile = del.genes, 
                          gisticScoresFile = scores.gistic)
# genome plot
gisticChromPlot(laml.gistic)

# gistic oncoplot
col <- c("red", "blue")
names(col) <- c('Amp', 'Del')
gisticOncoPlot(gistic = laml.gistic, colors = col)

```

## Waterfall plot

```{r}
# load the GenVisR package
library("GenVisR")
set.seed(426)

mutations.plot <- mutations.sig
colnames(mutations.plot)[c(1,2)]=c("Hugo_Symbol", "Tumor_Sample_Barcode")

waterfall(mutations.plot, 
          mainRecurCutoff = 0, 
          mainGrid = TRUE,
          plotMutBurden = TRUE, 
          fileType = "MAF", 
          coverageSpace = 90697072,
          plot_proportions = TRUE,
          proportions_type = "TvTi",
          mainDropMut = TRUE,
          rmvSilent = TRUE
          )
#clinData = gender,
#mainXlabel = TRUE,
```

## Analyze IRF mutations

```{r}
library(maftools)
library(gridExtra)
library(dplyr)

# Define the list of genes within the IRF pathway (suggested by Anja + combined with important IRF family members)
irf.genes <- c("TLR2", "TLR4", "TLR5", "MYD88", "RELA", "NFKB1", "TRAM1", "TICAM1", "TRAF6", "IRAK1", "IRAK4", "JAK2", "STAT1", "BCL6", "CD40", "IRF2BP2", "IRF8", "IRF4", "IRF1")
irf.genes <- c("IRF2BP2", "IRF8", "IRF4", "IRF1", "IRF3", "IRF7", "IRF9")

mutations <- read.table("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/mutation_summary-paired_and_unpaired.txt", sep = "\t", header = TRUE)

# extract IRF-familiy mutation data
mutations.irf_family <- unique(rbind(mutations[grep("^IRF", mutations$gene), ]))
mutations.irf_family$gene <- factor(mutations.irf_family$gene)
mutations.irf_family$effect <- factor(mutations.irf_family$effect)
mutations.irf_family.table <- table(mutations.irf_family$gene, mutations.irf_family$effect)
write.table(mutations.irf_family.table, "/Volumes/shung/projects/PMBCL_exomes/pathway_definitions/IRF/irf.mutation.table.txt", sep = "\t", quote = FALSE)

# extract mutation data for the complete IRF pathway
mutations.irf_pathway <- mutations[mutations$gene %in% irf.genes, ]
mutations.irf_pathway$gene <- factor(mutations.irf_pathway$gene)
mutations.irf_pathway$effect <- factor(mutations.irf_pathway$effect)
mutations.irf_pathway.table <- t(table(mutations.irf_pathway$gene, mutations.irf_pathway$effect))
write.table(mutations.irf_pathway, "/Volumes/shung/projects/PMBCL_exomes/pathway_definitions/IRF/mutations.irf_pathway.txt", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(mutations.irf_pathway.table, "/Volumes/shung/projects/PMBCL_exomes/pathway_definitions/IRF/mutations.irf_pathway.table.txt", sep = "\t", quote = FALSE)

# remove genes that are not mutated (not including UTR mutations)
irf.genes.not_mutated <- c("MYD88", "RELA", "TRAM1", "TICAM1", "IRAK1", "CD40")
irf.genes <- irf.genes[ !irf.genes %in% irf.genes.not_mutated ]

# read in MAF files to plot IRF-specific oncoplot and lollipop plots
mutations.maf <- read.maf("/Volumes/shung/projects/PMBCL_exomes/snv_indel_analysis/maf/mutation_summary-paired_and_unpaired_with_UTRs.maf")

lolli.plots <- vector('list', length(irf.genes))
for (i in seq_along(irf.genes)) {
  lolli.plots[[i]] <- lollipopPlot(maf = mutations.maf, 
                          gene = irf.genes[i], 
                          AACol = 'Protein_Change', 
                          showMutationRate = TRUE, 
                          domainLabelSize = 2, 
                          defaultYaxis = FALSE
                          )
}
do.call("grid.arrange", c(lolli.plots, ncol=1))

```

