---
title: "SNV Heatmap"
author: "Stacy Hung"
date: "March 26, 2019"
output: html_document
---

This script generates an oncoplot that orders mutations by pathology and for genes that have been selected accordingly:
  * Gene must be present in the target gene panel (n=217)
    * Gene is mutated in 10%+ of primary mediastinal
    * Gene is mutated in 10%+ of secondary mediastinal + non-mediastinal
    
Mutations have been filtered as follows:
  - somatic effect, excluding UTR mutations
  - filters used for SNV/indel analysis pipeline (default filters for t vs. blood and optimized filters for t vs. ME)
  
NB: Previously, genes were filtered based on the requirement of being mutated in 2+ cases; in this case, we do not want to apply this filter since we want a bigger picture across pathologies.

## Load packages

```{r}
library(VariantAnnotation)
library(GenVisR)
library(maftools)
library(gridExtra)
library(dplyr)
library(plyr)
```

## Get list of genes to use for oncoplot

```{r}
# get full list of mutations that will be potentially used for oncoplot (excludes GZ2229 only)
mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes_incl_GZ222-snvs_indels.maf", sep = "\t", header = TRUE)

# load target panel genes
genes.target <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/WES_data/target_genes.FINAL_217.txt", header = FALSE)
colnames(genes.target) <- c("gene")
genes.target <- as.vector(genes.target$gene)

# load metadata
clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.with_updated_mediastinal_and_age.txt", sep = "\t", header = TRUE)

# filter mutation dataset for target genes --> there are 151 genes mutated that are also present on the target panel
mutations.df <- filter(mutations.df, mutations.df$Hugo_Symbol %in% genes.target)
mutations.df$Hugo_Symbol <- factor(mutations.df$Hugo_Symbol)

# map on formal patient id mappings
patient_id.to.new_patient_id <- as.vector(clinical.data$Tumor_Sample_Barcode)
names(patient_id.to.new_patient_id) <- as.vector(clinical.data$tumor_id2)
mutations.df$Tumor_Sample_Barcode <- revalue(mutations.df$Tumor_Sample_Barcode, patient_id.to.new_patient_id)

# map on mediastrinal involvement (Not involved, secondary, primary)
patient.to.med_status <- as.vector(clinical.data$Mediastinal_involvement)
names(patient.to.med_status) <- as.vector(clinical.data$Tumor_Sample_Barcode)
mutations.df$med_status <- revalue(mutations.df$Tumor_Sample_Barcode, patient.to.med_status)

# calculate NUMBER OF CASES mutated in each GENE per mediastinal grouping
mutated_cases.by_medstat_and_gene <- unique(mutations.df[c("Hugo_Symbol", "Tumor_Sample_Barcode", "med_status")])
mutated.table <- as.data.frame(table(mutated_cases.by_medstat_and_gene$med_status,
                                     mutated_cases.by_medstat_and_gene$Hugo_Symbol))
colnames(mutated.table) <- c("Med_status", "gene", "num_mutated")

mutated.primary <- unique(filter(mutated_cases.by_medstat_and_gene, 
                                 mutated_cases.by_medstat_and_gene$med_status == "Primary"))
mutated.non_primary <- unique(filter(mutated_cases.by_medstat_and_gene, 
                                 mutated_cases.by_medstat_and_gene$med_status != "Primary"))

# get NUMBER OF CASES mutated per GROUP (to calculate normalized proportions)
NUM_PRIMARY <- length(unique(mutated.primary$Tumor_Sample_Barcode))
NUM_NONPRIMARY <- length(unique(mutated.non_primary$Tumor_Sample_Barcode))

# extract GENES mutated in each GROUP 
num_mutated.by_gene.primary <- subset(mutated.table, mutated.table$Med_status == "Primary" & mutated.table$num_mutated > 0)
num_mutated.by_gene.non_primary <- subset(mutated.table, mutated.table$Med_status != "Primary" & mutated.table$num_mutated > 0)

# calculate proportion of mutated cases
num_mutated.by_gene.primary$proportion <- num_mutated.by_gene.primary$num_mutated / NUM_PRIMARY * 100
num_mutated.by_gene.non_primary$proportion <- num_mutated.by_gene.non_primary$num_mutated / NUM_NONPRIMARY * 100

PERCENT_MUTATED = 10

# identify genes are present in 10%+ of primary mediastinal group, genes present in 10%+ of non-primary mediastinal group
num_mutated.med_primary.cutoff <- subset(num_mutated.by_gene.primary, num_mutated.by_gene.primary$proportion >= PERCENT_MUTATED)
num_mutated.med_nonprimary.cutoff <- subset(num_mutated.by_gene.non_primary, num_mutated.by_gene.non_primary$proportion >= PERCENT_MUTATED)
med_status_genes.10_plus <- unique(c(as.vector(num_mutated.med_primary.cutoff$gene),
                                               as.vector(num_mutated.med_nonprimary.cutoff$gene)))

write.table(med_status_genes.10_plus, "/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/med_status_genes.10_plus-v_incl_GZ222.txt", quote = FALSE, row.names = FALSE)

# create mutation matrix for filtered gene list
#mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.mutated_target_panel.maf", sep = "\t", header = TRUE, fill = TRUE)
mutations.df <- filter(mutations.df, mutations.df$Hugo_Symbol %in% med_status_genes.10_plus)
write.table(mutations.df, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes_incl_GZ222-snvs_indels.filtered_med_status_group.maf", sep = "\t", quote = FALSE, row.names = FALSE)

mutations.df$Hugo_Symbol <- factor(mutations.df$Hugo_Symbol)
mutations.df <- unique(mutations.df[c("Hugo_Symbol", "Tumor_Sample_Barcode")])
mutation.matrix <- table(mutations.df$Tumor_Sample_Barcode, mutations.df$Hugo_Symbol)
write.table(mutation.matrix, "/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/mutation_matrix-incl_GZ222-mediastinal_groups_10+_freq_mutated.txt", sep = "\t", quote = FALSE)
```

## Sort matrix

```{r}
# for each gene, calculate number of mutated cases; and number of pathology group-specific cases
#mut.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/mutation_matrix-mediastinal_groups_10+_freq_mutated.with_group_and_sums.txt", sep = "\t", header = TRUE)
mut.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/mutation_matrix_incl_GZ222-mediastinal_groups_10+_freq_mutated.with_group_and_sums.txt", sep = "\t", header = TRUE)

# sort by each gene in their current order

#mut.data.sorted <- mut.data[with(mut.data, order(mediastinal_involvement, -SOCS1, -B2M, -GNA13, -LRRN3, -CSMD3, -IRF2BP2, -CSF2RB, -ITPKB, -NCOR1, -ZNF217, -KCNQ3, -RELN, -ABCC6, -ARID1A, -C2CD3, -CD58, -EGR1, -IRF4, -LMF2, -PRKDC, -RC3H2, -SPEN, -STAT6, -WEE1, -ABCA13, -CISH, -MFHAS1, -MLL2, -NFKBIE, -SIAH2, -TNFAIP3, -ABCD2, -BRWD3, -DDX60, -DIAPH3, -GRIA4, -MAPK6, -MLL, -USP5, -XCR1, -ARHGAP29, -BTG1, -CD83, -CIITA, -DSC2, -DUSP2, -HDAC4, -HEXIM1, -HIST1H4C, -HIST1H4H, -IKBKB, -LTB, -NBPF6, -NF1, -NFATC3, -NFKBIA, -NONO, -NSD1, -PCBP1, -PIK3C2G, -POU4F3, -PRDM16, -RNF217, -SRSF2, -TBC1D4, -TNIP1, -WDFY3, -ZFP36L1, -ZNF423)), ]

mut.data.sorted <- mut.data[with(mut.data, order(mediastinal_involvement, -SOCS1, -B2M, -GNA13, -LRRN3, -CSMD3, -IRF2BP2, -CSF2RB, -ITPKB, -NCOR1, -NFKBIE, -ZNF217, -KCNQ3, -MLL, -RELN, -ABCC6, -ARID1A, -C2CD3, -CD58, -EGR1, -IRF4, -LMF2, -PRKDC, -RC3H2, -SPEN, -STAT6, -WEE1, -ABCA13, -CISH, -MFHAS1, -MLL2, -SIAH2, -TNFAIP3, -ZFP36L1, -ABCD2, -BRWD3, -DDX60, -DIAPH3, -GRIA4, -MAPK6, -USP5, -XCR1, -ARHGAP29, -BTG1, -CD83, -CIITA, -DDX3X, -DSC2, -DUSP2, -HDAC4, -HEXIM1, -HIST1H4C, -HIST1H4H, -IKBKB, -LTB, -NBPF6, -NF1, -NFATC3, -NFKBIA, -NONO, -NSD1, -PCBP1, -PIK3C2G, -POU4F3, -PRDM16, -RNF217, -SRSF2, -TBC1D4, -TNIP1, -WDFY3, -ZNF423)), ]

patients.ordered <- as.vector(mut.data.sorted$patient)
write.table(patients.ordered, "/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/patients.ordered.CURRENT.txt", sep = "\t", quote = FALSE)

mut.oncoplot <- t(mut.data.sorted)
genes.ordered <- as.vector(row.names(mut.oncoplot))
genes.ordered <- genes.ordered[ genes.ordered != "patient"]
genes.ordered <- genes.ordered[ genes.ordered != "mediastinal_involvement"]
write.table(genes.ordered, "/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/genes.ordered.CURRENT.txt", sep = "\t", quote = FALSE)

write.table(mut.oncoplot, "/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/mutation_matrix_incl_GZ222-mediastinal_groups_10+_freq_mutated.oncoplot.txt", sep = "\t", quote = FALSE)

```

## Mediastinal involvement-specific oncoplots

```{r}
library(maftools)

clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.with_updated_mediastinal_and_age_and_9p24.txt", sep = "\t", header = TRUE)

#mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.filtered_med_status_group.maf", clinicalData = clinical.data)
#mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/gzl_exomes-snvs_indels.filtered_med_status_group.with_REMOVE_gene.maf", clinicalData = clinical.data)

# altered version with a "REMOVE" gene so that patients with no mutations are still represented
mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes_incl_GZ222-snvs_indels.filtered_med_status_group.with_REMOVE_gene.maf", clinicalData = clinical.data)

genes.ordered <- read.table("/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/genes.ordered.CURRENT.txt")
genes.ordered <- as.vector(genes.ordered$x)
patients.ordered <- read.table("/Volumes/shung/projects/gzl_exomes/data/mediastinal_group_visualization/patients.ordered.CURRENT.txt")
patients.ordered <- as.vector(patients.ordered$x)

# colours for annotation tracks
annot.colors <- list(Gender = c("Female" = "palevioletred1", "Male" = "royalblue1"),
                     EBER = c("Negative" = "black", "Positive" = "red"),
                     Mediastinal = c("Not_involved" = "grey", "Secondary" = "plum2", "Primary" = "purple"),
                     Age_yrs = c("<_45" = "bisque", "45_<_age_<_60"="navajowhite3", ">=_60" = "bisque4"),
                     Path_group = c("0_=_cHL" = "lightsalmon",
                                   "1_=_cHL-interm" = "khaki1",
                                   "2_=_LBCL-interm" = "lightgreen", 
                                   "3_=_LBCL" = "steelblue1")
               )

#                     GISTIC_9p24 = c("no_change" = "grey", 
#                                     "low-level_amp" = "#E6C8C8", 
#                                     "high-level_amp" = "firebrick3"),
#                     FISH_9p24 = c("NORMAL" = "grey" , "GAIN" = "indianred1", "AMP" = "firebrick3"),


# show patients ordered by group, and mutations ordered by frequency
# include "REMOVE" gene since the removeNonMutated parameter doesn't work (i.e. doesn't show non-mutated case when FALSE)
genes.ordered <- c(genes.ordered, "REMOVE")

oncoplot(maf = mutations.maf, SampleNamefontSize = 0.8, 
         removeNonMutated = FALSE,
         sampleOrder = patients.ordered,
         genes = genes.ordered, 
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 1, annotationFontSize = 1,
         clinicalFeatures = c("Mediastinal_involvement", "EBER", "Gender", 
                              "Path_group", "Age_category"),
         annotationColor = annot.colors, 
         showTumorSampleBarcodes = TRUE,
         drawColBar = FALSE)

#                              "GISTIC_9p24", "FISH_9p24",  

# version with large patient font
oncoplot(maf = mutations.maf, SampleNamefontSize = 0.75,
         removeNonMutated = FALSE,
         sampleOrder = patients.ordered, 
         genes = genes.ordered, 
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 0, annotationFontSize = 0,
         showTumorSampleBarcodes = TRUE,
         drawColBar = FALSE)

```

