---
title: "Oncoplot for MutSig genes"
author: "Stacy Hung"
date: "October , 2019"
output: html_document
---

This script generates an oncoplot that orders mutations by pathology and for genes that have been predicted to be significant according to MutSig
  
Mutations have been filtered as follows:
  - somatic effect, excluding UTR mutations
  - filters used for SNV/indel analysis pipeline (default filters for t vs. blood and optimized filters for t vs. ME)
  
NB: Previously, genes were filtered based on the requirement of being mutated in 2+ cases; in this case, we do not want to apply this filter since we want a bigger picture across pathologies.

## Load packages

```{r}
library(maftools)
library(gridExtra)
library(dplyr)
library(plyr)
```

## Path group-specific oncoplots

```{r}
clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.with_updated_mediastinal_and_age.txt", sep = "\t", header = TRUE)
clinical.data$Path_group <- revalue(clinical.data$Path_group, 
                                          c("0 = cHL"="0_=_cHL",
                                            "1 = cHL-interm"="1_=_cHL-interm",
                                            "2 = LBCL-interm"="2_=_LBCL-interm",
                                            "3 = LBCL"="3_=_LBCL"))
# extract tumor id mappings
id_mappings <- clinical.data[,c("Tumor_Sample_Barcode", "tumor_id2")]

# read in mutations as dataframe (so that we can map tumor ids)
mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/maf/snvs_indels.full_cohort_minus_GZ229.default_and_optimized.target_genes.no_UTRs.maf", sep = "\t", header = TRUE, fill = TRUE, stringsAsFactors = TRUE)

# add mappable tumor code (to be consistent with clinical data)
mutations.df$tumor_id2 <- mutations.df$Tumor_Sample_Barcode
mutations.df$Tumor_Sample_Barcode <- NULL
mutations.df <- merge(mutations.df, id_mappings, by = "tumor_id2")

# write out new maf
write.table(mutations.df, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/maf/snvs_indels.full_cohort_minus_GZ229.default_and_optimized.target_genes.no_UTRs.for_oncoplot.maf", sep = "\t", quote = FALSE, row.names = FALSE)

mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/maf/snvs_indels.full_cohort_minus_GZ229.default_and_optimized.target_genes.no_UTRs.for_oncoplot.maf", clinicalData = clinical.data)

# colours for annotation tracks
annot.colors <- list(EBER = c("Negative" = "black", "Positive" = "red"),
                     Mediastinal = c("Not_involved" = "grey", "Secondary" = "plum2", "Primary" = "purple"),
                     Path_group = c("0_=_cHL" = "lightsalmon",
                                   "1_=_cHL-interm" = "khaki1",
                                   "2_=_LBCL-interm" = "lightgreen", 
                                   "3_=_LBCL" = "steelblue1")
               )

oncoplot(maf = mutations.maf, SampleNamefontSize = 0.65, top = 36, 
         removeNonMutated = FALSE,
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 1, annotationFontSize = 1,
         clinicalFeatures = c("Path_group", "EBER", "Mediastinal_involvement"),
         annotationColor = annot.colors, 
         showTumorSampleBarcodes = TRUE,
         drawColBar = FALSE, writeMatrix = TRUE)

# version with large patient font
oncoplot(maf = mutations.maf, SampleNamefontSize = 0.75,
         removeNonMutated = FALSE,
         sampleOrder = patients.ordered, 
         genes = genes.ordered, 
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 0, annotationFontSize = 0,
         showTumorSampleBarcodes = TRUE,
         drawColBar = FALSE)

```

