---
title: "SNV Heatmap"
author: "Stacy Hung"
date: "March 26, 2019"
output: html_document
---

This script generates an oncoplot that orders mutations by EBER status and for genes that have been selected accordingly:
  * Gene must be present in the target gene panel (n=217)
    * Gene is mutated in 10%+ of EBV+ cases
    * Gene is mutated in 10%+ of EBV- cases
    
Mutations have been filtered as follows:
  - somatic effect, excluding UTR mutations
  - filters used for SNV/indel analysis pipeline (default filters for t vs. blood and optimized filters for t vs. ME)
  - GZ222 and GZ229 were excluded from the analysis as they are considered failures (tumor contamination)
  
NB: Previously, genes were filtered based on the requirement of being mutated in 2+ cases; in this case, we do not want to apply this filter since we want a bigger picture across pathologies.


## Load packages

```{r}
# NB: if you need to install this package, install it using devtools from the github source!!!
# The bioconductor version of GenVisR is buggy (not compatible with required libraries)
# Installing GenVisR may also require the installation of the VariantAnnotation pacakge (from bioconductor)
source("https://bioconductor.org/biocLite.R")
biocLite("VariantAnnotation")
library("VariantAnnotation")

install.packages("devtools")
library(devtools)
install_github("griffithlab/GenVisR")
library("GenVisR")

source("https://bioconductor.org/biocLite.R")
biocLite("maftools")
library(maftools)

library(gridExtra)
```

## Get list of genes to use for oncoplot
```{r}
library(dplyr)
library(plyr)

# get full list of mutations that will be potentially used for oncoplot (excludes GZ2229 only)
mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes_incl_GZ222-snvs_indels.maf", sep = "\t", header = TRUE)

# get full list of mutations that will be potentially used for oncoplot
#mutations.df <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes-snvs_indels.maf", sep = "\t", header = TRUE)

# load target panel genes
genes.target <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/WES_data/target_genes.FINAL_217.txt", header = FALSE)
colnames(genes.target) <- c("gene")
genes.target <- as.vector(genes.target$gene)

# load metadata
clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.with_updated_mediastinal_and_age.txt", sep = "\t", header = TRUE)

# filter mutation dataset for target genes --> there are 151 genes mutated that are also present on the target panel
mutations.df <- filter(mutations.df, mutations.df$Hugo_Symbol %in% genes.target)
mutations.df$Hugo_Symbol <- factor(mutations.df$Hugo_Symbol)

# map on formal patient id mappings
patient_id.to.new_patient_id <- as.vector(clinical.data$Tumor_Sample_Barcode)
names(patient_id.to.new_patient_id) <- as.vector(clinical.data$tumor_id2)
mutations.df$Tumor_Sample_Barcode <- revalue(mutations.df$Tumor_Sample_Barcode, patient_id.to.new_patient_id)

# map on EBER status (postive / negative)
patient.to.EBER <- as.vector(clinical.data$EBER)
names(patient.to.EBER) <- as.vector(clinical.data$Tumor_Sample_Barcode)
mutations.df$EBER <- revalue(mutations.df$Tumor_Sample_Barcode, patient.to.EBER)

# calculate NUMBER OF CASES mutated in each GENE per EBER grouping
mutated_cases.by_eber_and_gene <- unique(mutations.df[c("Hugo_Symbol", "Tumor_Sample_Barcode", "EBER")])
mutated.table <- as.data.frame(table(mutated_cases.by_eber_and_gene$EBER, 
                                     mutated_cases.by_eber_and_gene$Hugo_Symbol))
colnames(mutated.table) <- c("EBER_status", "gene", "num_mutated")

mutated.EBER_pos <- unique(filter(mutated_cases.by_eber_and_gene, mutated_cases.by_eber_and_gene$EBER == "Positive"))
mutated.EBER_neg <- unique(filter(mutated_cases.by_eber_and_gene, mutated_cases.by_eber_and_gene$EBER == "Negative"))

# get NUMBER OF CASES mutated per GROUP (to calculate normalized proportions)
NUM_EBER_POS <- length(unique(mutated.EBER_pos$Tumor_Sample_Barcode))
NUM_EBER_NEG <- length(unique(mutated.EBER_neg$Tumor_Sample_Barcode))
NUM_ALL_CASES <- length(unique(mutated_cases.by_eber_and_gene$Tumor_Sample_Barcode))

# extract GENES mutated in each GROUP 
num_mutated.by_EBER_pos <- subset(mutated.table, mutated.table$EBER_status == "Positive" & mutated.table$num_mutated > 0)
num_mutated.by_EBER_neg <- subset(mutated.table, mutated.table$EBER_status == "Negative" & mutated.table$num_mutated > 0)
num_mutated.all_cases <- subset(mutated.table, mutated.table$num_mutated > 0)

# calculate proportion of mutated cases
num_mutated.by_EBER_pos$proportion <- num_mutated.by_EBER_pos$num_mutated / NUM_EBER_POS*100
num_mutated.by_EBER_neg$proportion <- num_mutated.by_EBER_neg$num_mutated / NUM_EBER_NEG*100
num_mutated.all_cases$proportion <- num_mutated.all_cases$num_mutated / NUM_ALL_CASES*100

PERCENT_MUTATED = 10
MIN_CASES_MUTATED = 2

# CRITERIA FOR GENE SELECTION:

# CURRENT CRITERIA: genes that are mutated in 2+ cases in EBV+ cases *OR* mutated in >= 10% of ALL cases
# Previous criteria: genes are mutated in >= 10% for each group
num_mutated.EBER_pos.cutoff <- subset(num_mutated.by_EBER_pos, num_mutated.by_EBER_pos$num_mutated >= MIN_CASES_MUTATED)
num_mutated.all_cases.cutoff <- subset(num_mutated.all_cases, num_mutated.all_cases$proportion >= PERCENT_MUTATED)

#eber_genes.10_plus <- unique(c(as.vector(num_mutated.EBER_pos.cutoff$gene), as.vector(num_mutated.EBER_neg.cutoff$gene)))
eber_genes.10_plus <- unique(c(as.vector(num_mutated.EBER_pos.cutoff$gene), as.vector(num_mutated.all_cases.cutoff$gene)))

write.table(eber_genes.10_plus, "/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/eber_genes.10_plus_all_cases.2+_cases_EBV_pos-v_incl_GZ222.txt", quote = FALSE, row.names = FALSE)

# create mutation matrix for filtered gene list
mutations.df <- filter(mutations.df, mutations.df$Hugo_Symbol %in% eber_genes.10_plus)
write.table(mutations.df, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes_incl_GZ222-snvs_indels.filtered_eber_status_v2.maf", sep = "\t", quote = FALSE, row.names = FALSE)

mutations.df$Hugo_Symbol <- factor(mutations.df$Hugo_Symbol)
mutations.df <- unique(mutations.df[c("Hugo_Symbol", "Tumor_Sample_Barcode")])
mutation.matrix <- table(mutations.df$Tumor_Sample_Barcode, mutations.df$Hugo_Symbol)
write.table(mutation.matrix, "/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/mutation_matrix-incl_GZ222-eber_groups_10+_freq_mutated.2+_cases_EBER_pos.txt", sep = "\t", quote = FALSE)

```

## Sort matrix

```{r}
# manual steps that are easier in Excel:
# map EBER status for each patient
# sort by EBER status (negative, positive)
# transpose, calculate sums across cohort, and each group
# sort by total sum (decreassing), sum of patients in EBER+ (increasing), and sum of patients in EBER- (increasing)
# --> this ensures that genes unique mutated in EBER+ show up at the bottom of the plot

# for each gene, calculate number of mutated cases; and number of pathology group-specific cases
#mut.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/mutation_matrix-incl_GZ222-eber_groups_10+_freq_mutated.sorted_by_grp_sums.txt", sep = "\t", header = TRUE)
mut.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/mutation_matrix-incl_GZ222-eber_groups_10+_freq_mutated.2+_cases_EBER_pos.sorted_by_group_sums.txt", sep = "\t", header = TRUE)

# sort by each gene in their current order
#mut.data.sorted <- mut.data[with(mut.data, order(EBER_status, -SOCS1, -B2M, -LRRN3, -GNA13, -ITPKB, -NCOR1, -NFKBIE, -CSF2RB, -CSMD3, -IRF2BP2, -ZNF217, -ABCA13, -ABCC6, -ARID1A, -EGR1, -KCNQ3, -MFHAS1, -MLL2, -RC3H2, -SIAH2, -SPEN, -STAT6, -TNFAIP3, -WEE1, -ZFP36L1, -C2CD3, -CISH, -IRF4, -LMF2, -MLL, -PRKDC, -RELN, -CD58, -AMOT, -ASCC3, -BARX2, -DDX60, -EHMT2, -GPR137C, -GRIN3A, -HIST1H4C, -MAPK6, -MCL1, -NLRP4, -NOTCH2, -POM121L12, -RIBC2, -SEMA3E, -XCR1, -ACTB, -ATP13A4, -DHX58, -STAT3, -HIST1H1D, -NOTCH1, -RGS1, -SGK1, -TAF1)), ]

mut.data.sorted <- mut.data[with(mut.data, order(EBER_status, -SOCS1, -B2M, -LRRN3, -GNA13, -ITPKB, -NCOR1, -NFKBIE, -CSF2RB, -CSMD3, -IRF2BP2, -ZNF217, -ABCA13, -ABCC6, -ARID1A, -EGR1, -KCNQ3, -MFHAS1, -MLL2, -RC3H2, -SIAH2, -SPEN, -STAT6, -TNFAIP3, -WEE1, -ZFP36L1, -CD58, -ACTB, -ATP13A4, -DHX58, -STAT3)), ]

patients.ordered <- as.vector(mut.data.sorted$patient)
write.table(patients.ordered, "/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/patients.ordered.CURRENT.txt", sep = "\t", quote = FALSE)

mut.oncoplot <- t(mut.data.sorted)
genes.ordered <- as.vector(row.names(mut.oncoplot))
genes.ordered <- genes.ordered[ genes.ordered != "patient"]
genes.ordered <- genes.ordered[ genes.ordered != "EBER_status"]
write.table(genes.ordered, "/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/genes.ordered.CURRENT.txt", sep = "\t", quote = FALSE)

write.table(mut.oncoplot, "/Volumes/shung/projects/gzl_exomes/data/EBV_visualization/mutation_matrix-incl_GZ222-eber_groups_10+_freq_mutated.2+_cases_EBER_pos.oncoplot.txt", sep = "\t", quote = FALSE)
```

## EBV-specific oncoplots

```{r}
# to get latest version of maftools, install from github
library("devtools")
install_github(repo = "PoisonAlien/maftools")
library(maftools)

clinical.data <- read.table("/Volumes/shung/projects/gzl_exomes/data/GZ_WES_cases-clinical_data.with_updated_mediastinal_and_age.txt", sep = "\t", header = TRUE)

#mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes_incl_GZ222-snvs_indels.filtered_eber_status.maf", clinicalData = clinical.data)
mutations.maf <- read.maf("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/snv_indel_datasets/maf/gzl_exomes_incl_GZ222-snvs_indels.filtered_eber_status_v2.maf", clinicalData = clinical.data)

# colours for annotation tracks
annot.colors <- list(Gender = c("Female" = "palevioletred1", "Male" = "royalblue1"),
                     EBER = c("Negative" = "black", "Positive" = "red"),
                     Mediastinal = c("Not_involved" = "grey", "Secondary" = "plum2", "Primary" = "purple"),
                     Age_yrs = c("<_45" = "bisque", "45_<_age_<_60"="navajowhite3", ">=_60" = "bisque4"),
                     Path_group = c("0_=_cHL" = "lightsalmon",
                                   "1_=_cHL-interm" = "khaki1",
                                   "2_=_LBCL-interm" = "lightgreen", 
                                   "3_=_LBCL" = "steelblue1")
               )

# plot by EBV+ and EBV- sorted cohorts
# list of genes that are mutated in at least 10% of EBV+ and EBV- cases, sorted by gene

oncoplot(maf = mutations.maf, SampleNamefontSize = 0.7,
         removeNonMutated = FALSE,
         sampleOrder = patients.ordered, 
         genes = genes.ordered, 
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 1, annotationFontSize = 1,
         clinicalFeatures = c("EBER", "Path_group", "Mediastinal_involvement", "Gender", "Age_category"),
         annotationColor = annot.colors, 
         showTumorSampleBarcodes = TRUE,
         drawColBar = FALSE)

# version with large patient font
oncoplot(maf = mutations.maf, SampleNamefontSize = 0.75,
         removeNonMutated = FALSE,
         sampleOrder = patients.ordered, 
         genes = genes.ordered, 
         keepGeneOrder = TRUE, fontSize = 0.6, titleFontSize = 0, legendFontSize = 0, annotationFontSize = 0,
         showTumorSampleBarcodes = TRUE,
         drawColBar = FALSE)

```

