---
title: "MutationalPatterns for GZL exomes"
author: "Stacy Hung"
date: "April 19, 2019"
output: html_document
---

This script analyzes mutations associated with a given phenotype and COSMIC signature (e.g. what is the signficance of mutations attributable to AID/APOBEC activity and EBV status?).

COSMIC signature categories: 
(1) Aging - #1
(2) AID/APOBEC activity - #2, 13
(3) Defective DNA mismatch repair - #3, 6, 15, 20, 26
(4) Polymerase-n-mediated somatic hypermutation - #9
(5) All other signatures

Phenotypic groups:
(1) EBV status (+/-)
(2) Mediastinal status (+/-)
(3) Path group (1/2 or 0/3)
(3) Age group (<= 45, 45-60, >= 60)


Part 2: The extent to which mutations are associated with a given gene and COSMIC signature is also assessed (i.e. mutation in gene X and present in signature Y?)

## Load data and set variables

```{r}
plot_data <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/GZ-optimal_contributions-COSMIC.normalized.grouped.all_phenotypes.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

SIG_AID = "AID / APOBEC activity (2, 13)"
SIG_DDMR = "Defective DNA mismatch repair (3, 6, 15, 20, 26)"
SIG_HYPER = "Polymerase-n-mediated somatic hypermutation (9)"
SIG_AGING = "Aging (1)"
SIG_OTHERS = "All other COSMIC signatures"
SIG_POLE = ""

EBV_POS = "Positive"
EBV_NEG = "Negative"

MED_PRIMARY = 1
MED_NOT_PRIMARY = 0

PATH_GRP_1_2 = 1
PATH_GRP_0_3 = 0

AGE_45_UNDER = "<= 45"
AGE_45_TO_60 = "45 < age < 60"
AGE_60_PLUS = ">= 60"

contingency_table <- array(dim = c(10,9))
```

## #########################################################################
## Chi-square tests for EBV status vs. different COSMIC signatures
## #########################################################################

## Calculate enrichment of the AID/APOBEC signature in EBV vs. EBV- groups

```{r}
library(dplyr)

EBV_pos.AID_sig.data <- filter(plot_data, EBER_status == EBV_POS & signature_group == SIG_AID)
EBV_pos.not_AID_sig.data <- filter(plot_data, EBER_status == EBV_POS & !(signature_group == SIG_AID))
EBV_neg.AID_sig.data <- filter(plot_data, EBER_status == EBV_NEG & signature_group == SIG_AID)
EBV_neg.not_AID_sig.data <- filter(plot_data, EBER_status == EBV_NEG & !(signature_group == SIG_AID))

EBV_pos.AID_sig.num <- sum(EBV_pos.AID_sig.data$contribution) # 17
EBV_pos.not_AID_sig.num <- sum(EBV_pos.not_AID_sig.data$contribution) # 633
EBV_neg.AID_sig.num <- sum(EBV_neg.AID_sig.data$contribution) # 267
EBV_neg.not_AID_sig.num <- sum(EBV_neg.not_AID_sig.data$contribution) # 5158

#EBV_pos.total <- EBV_pos.AID_sig.num + EBV_pos.not_AID_sig.num
#EBV_neg.total <- EBV_neg.AID_sig.num + EBV_neg.not_AID_sig.num

# chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.AID_sig.num, EBV_neg.AID_sig.num)
sam[2, ] <- c(EBV_pos.not_AID_sig.num, EBV_neg.not_AID_sig.num)

contingency_table[1:2,1:2] <- sam

# Chi-Square test to assess if AID/APOBEC in EBV+ cases is more different than expected
pvalue <- chisq.test(sam)$p.value # 0.009979839; previously 0.01326989 (when using EBV_pos.total and EBV_neg.total)
```

## Calculate enrichment of defective DNA mismatch repair signature in EBV+ vs. EBV- groups

```{r}
library(dplyr)

EBV_pos.DDMR_sig.data <- filter(plot_data, EBER_status == EBV_POS & signature_group == SIG_DDMR)
EBV_pos.not_DDMR_sig.data <- filter(plot_data, EBER_status == EBV_POS & !(signature_group == SIG_DDMR))
EBV_neg.DDMR_sig.data <- filter(plot_data, EBER_status == EBV_NEG & signature_group == SIG_DDMR)
EBV_neg.not_DDMR_sig.data <- filter(plot_data, EBER_status == EBV_NEG & !(signature_group == SIG_DDMR))

EBV_pos.DDMR_sig.num <- sum(EBV_pos.DDMR_sig.data$contribution) # 191.3075
EBV_pos.not_DDMR_sig.num <- sum(EBV_pos.not_DDMR_sig.data$contribution) # 458.9504
EBV_neg.DDMR_sig.num <- sum(EBV_neg.DDMR_sig.data$contribution) # 1458.566
EBV_neg.not_DDMR_sig.num <- sum(EBV_neg.not_DDMR_sig.data$contribution) # 3966.872

#EBV_pos.total <- EBV_pos.DDMR_sig.num + EBV_pos.not_DDMR_sig.num
#EBV_neg.total <- EBV_neg.DDMR_sig.num + EBV_neg.not_DDMR_sig.num

# chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.DDMR_sig.num, EBV_neg.DDMR_sig.num)
sam[2, ] <- c(EBV_pos.not_DDMR_sig.num, EBV_neg.not_DDMR_sig.num)

contingency_table[3:4,1:2] <- sam
  
# Chi-Square test to assess if DDMR in EBV+ cases is more different than expected
pvalue <- chisq.test(sam)$p.value # 0.1843208; previously 0.3233782 (when using EBV_pos.total and EBV_neg.total)
```

## Calculate enrichment of hypermutation signature in EBV+ vs. EBV- groups

```{r}
EBV_pos.hyper_sig.num <- sum((filter(plot_data, EBER_status == EBV_POS & signature_group == SIG_HYPER))$contribution)
EBV_pos.not_hyper_sig.num <- sum((filter(plot_data, EBER_status == EBV_POS & !(signature_group == SIG_HYPER)))$contribution)
EBV_neg.hyper_sig.num <- sum((filter(plot_data, EBER_status == EBV_NEG & signature_group == SIG_HYPER))$contribution)
EBV_neg.not_hyper_sig.num <- sum((filter(plot_data, EBER_status == EBV_NEG & !(signature_group == SIG_HYPER)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.hyper_sig.num, EBV_neg.hyper_sig.num) # 25, 139
sam[2, ] <- c(EBV_pos.not_hyper_sig.num, EBV_neg.not_hyper_sig.num) # 625, 5286

contingency_table[5:6,1:2] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.07539388
```

## Calculate enrichment of aging signature in EBV+ vs. EBV- groups

```{r}
EBV_pos.aging_sig.num <- sum((filter(plot_data, EBER_status == EBV_POS & signature_group == SIG_AGING))$contribution)
EBV_pos.not_aging_sig.num <- sum((filter(plot_data, EBER_status == EBV_POS & !(signature_group == SIG_AGING)))$contribution)
EBV_neg.aging_sig.num <- sum((filter(plot_data, EBER_status == EBV_NEG & signature_group == SIG_AGING))$contribution)
EBV_neg.not_aging_sig.num <- sum((filter(plot_data, EBER_status == EBV_NEG & !(signature_group == SIG_AGING)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.aging_sig.num, EBV_neg.aging_sig.num) # 155, 1626
sam[2, ] <- c(EBV_pos.not_aging_sig.num, EBV_neg.not_aging_sig.num) # 495, 3800

contingency_table[7:8,1:2] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.001396633
```

## Calculate enrichment of all other COSMIC signatures in EBV+ vs. EBV- groups

```{r}
EBV_pos.others_sig.num <- sum((filter(plot_data, EBER_status == EBV_POS & signature_group == SIG_OTHERS))$contribution)
EBV_pos.not_others_sig.num <- sum((filter(plot_data, EBER_status == EBV_POS & !(signature_group == SIG_OTHERS)))$contribution)
EBV_neg.others_sig.num <- sum((filter(plot_data, EBER_status == EBV_NEG & signature_group == SIG_OTHERS))$contribution)
EBV_neg.not_others_sig.num <- sum((filter(plot_data, EBER_status == EBV_NEG & !(signature_group == SIG_OTHERS)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.others_sig.num, EBV_neg.others_sig.num) # 262, 1935
sam[2, ] <- c(EBV_pos.not_others_sig.num, EBV_neg.not_others_sig.num) # 388, 3490

contingency_table[9:10,1:2] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.02224143
```


## #########################################################################
## Chi-square tests for mediastinal status vs. different COSMIC signatures
## #########################################################################

## Test association between mediastinal status and AID/APOBEC activity signature

```{r}
Med_1.AID_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & signature_group == SIG_AID))$contribution)
Med_1.not_AID_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & !(signature_group == SIG_AID)))$contribution)
Med_not_1.AID_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & signature_group == SIG_AID))$contribution)
Med_not_1.not_AID_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & !(signature_group == SIG_AID)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(Med_1.AID_sig.num, Med_not_1.AID_sig.num) # 167, 116
sam[2, ] <- c(Med_1.not_AID_sig.num, Med_not_1.not_AID_sig.num) # 4353, 1438

contingency_table[1:2,3:4] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 1.725684e-09
```

## Test association between mediastinal status and DNA mismatch repair signature

```{r}
Med_1.DDMR_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & signature_group == SIG_DDMR))$contribution)
Med_1.not_DDMR_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & !(signature_group == SIG_DDMR)))$contribution)
Med_not_1.DDMR_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & signature_group == SIG_DDMR))$contribution)
Med_not_1.not_DDMR_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & !(signature_group == SIG_DDMR)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(Med_1.DDMR_sig.num, Med_not_1.DDMR_sig.num) # 1207, 443
sam[2, ] <- c(Med_1.not_DDMR_sig.num, Med_not_1.not_DDMR_sig.num) # 3314, 1112

contingency_table[3:4,3:4] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.1837358
```

## Test association between mediastinal status and hypermutation signature

```{r}
Med_1.hyper_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & signature_group == SIG_HYPER))$contribution)
Med_1.not_hyper_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & !(signature_group == SIG_HYPER)))$contribution)
Med_not_1.hyper_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & signature_group == SIG_HYPER))$contribution)
Med_not_1.not_hyper_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & !(signature_group == SIG_HYPER)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(Med_1.hyper_sig.num, Med_not_1.hyper_sig.num) # 139, 25
sam[2, ] <- c(Med_1.not_hyper_sig.num, Med_not_1.not_hyper_sig.num) # 4381, 1530 

contingency_table[5:6,3:4] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.002753807
```

## Test association between mediastinal status and aging signature

```{r}
Med_1.aging_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & signature_group == SIG_AGING))$contribution)
Med_1.not_aging_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & !(signature_group == SIG_AGING)))$contribution)
Med_not_1.aging_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & signature_group == SIG_AGING))$contribution)
Med_not_1.not_aging_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & !(signature_group == SIG_AGING)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(Med_1.aging_sig.num, Med_not_1.aging_sig.num) # 1423, 357
sam[2, ] <- c(Med_1.not_aging_sig.num, Med_not_1.not_aging_sig.num) # 3097, 1198 

contingency_table[7:8,3:4] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 2.376584e-10
```

## Test association between mediastinal status and all other COSMIC signatures

```{r}
Med_1.others_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & signature_group == SIG_OTHERS))$contribution)
Med_1.not_others_sig.num <- sum((filter(plot_data, mediastinal_status == MED_PRIMARY & !(signature_group == SIG_OTHERS)))$contribution)
Med_not_1.others_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & signature_group == SIG_OTHERS))$contribution)
Med_not_1.not_others_sig.num <- sum((filter(plot_data, mediastinal_status == MED_NOT_PRIMARY & !(signature_group == SIG_OTHERS)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(Med_1.others_sig.num, Med_not_1.others_sig.num) # 1583, 614
sam[2, ] <- c(Med_1.not_others_sig.num, Med_not_1.not_others_sig.num) # 2937, 941

contingency_table[9:10,3:4] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.001892591
```



## #########################################################################
## Chi-square tests for pathology grouping vs. different COSMIC signatures
## #########################################################################

## Test association between pathology group and the AID/APOBEC activity signature

```{r}
path_1_2.AID_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & signature_group == SIG_AID))$contribution)
path_1_2.not_AID_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & !(signature_group == SIG_AID)))$contribution)
path_0_3.AID_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & signature_group == SIG_AID))$contribution)
path_0_3.not_AID_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & !(signature_group == SIG_AID)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(path_1_2.AID_sig.num, path_0_3.AID_sig.num) # 269, 15
sam[2, ] <- c(path_1_2.not_AID_sig.num, path_0_3.not_AID_sig.num) # 5081, 710

contingency_table[1:2,5:6] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.0005758003
```

## Test association between pathology group and the DNA mismatch repair signature

```{r}
path_1_2.DDMR_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & signature_group == SIG_DDMR))$contribution)
path_1_2.not_DDMR_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & !(signature_group == SIG_DDMR)))$contribution)
path_0_3.DDMR_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & signature_group == SIG_DDMR))$contribution)
path_0_3.not_DDMR_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & !(signature_group == SIG_DDMR)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(path_1_2.DDMR_sig.num, path_0_3.DDMR_sig.num) # 1427, 223
sam[2, ] <- c(path_1_2.not_DDMR_sig.num, path_0_3.not_DDMR_sig.num) # 3924, 502

contingency_table[3:4,5:6] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.02151707
```

## Test association between pathology group and the hypermutation signature

```{r}
path_1_2.hyper_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & signature_group == SIG_HYPER))$contribution)
path_1_2.not_hyper_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & !(signature_group == SIG_HYPER)))$contribution)
path_0_3.hyper_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & signature_group == SIG_HYPER))$contribution)
path_0_3.not_hyper_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & !(signature_group == SIG_HYPER)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(path_1_2.hyper_sig.num, path_0_3.hyper_sig.num) # 146, 18
sam[2, ] <- c(path_1_2.not_hyper_sig.num, path_0_3.not_hyper_sig.num) # 5204, 708

contingency_table[5:6,5:6] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.7479971
```

## Test association between pathology group and the aging signature

```{r}
path_1_2.aging_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & signature_group == SIG_AGING))$contribution)
path_1_2.not_aging_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & !(signature_group == SIG_AGING)))$contribution)
path_0_3.aging_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & signature_group == SIG_AGING))$contribution)
path_0_3.not_aging_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & !(signature_group == SIG_AGING)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(path_1_2.aging_sig.num, path_0_3.aging_sig.num) # 1582, 199
sam[2, ] <- c(path_1_2.not_aging_sig.num, path_0_3.not_aging_sig.num) # 3768, 527

contingency_table[7:8,5:6] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.2483097
```

## Test association between pathology group and all other COSMIC signatures

```{r}
path_1_2.others_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & signature_group == SIG_OTHERS))$contribution)
path_1_2.not_others_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_1_2 & !(signature_group == SIG_OTHERS)))$contribution)
path_0_3.others_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & signature_group == SIG_OTHERS))$contribution)
path_0_3.not_others_sig.num <- sum((filter(plot_data, path_group == PATH_GRP_0_3 & !(signature_group == SIG_OTHERS)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(path_1_2.others_sig.num, path_0_3.others_sig.num) # 1927, 270
sam[2, ] <- c(path_1_2.not_others_sig.num, path_0_3.not_others_sig.num) # 3424, 455

contingency_table[9:10,5:6] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 0.5305204
```


## #########################################################################
## Chi-square tests for age groups vs. different COSMIC signatures
## #########################################################################

## Test association between age groups and the AID/APOBEC activity signature

```{r}
age_45_under.AID_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & signature_group == SIG_AID))$contribution)
age_45_to_60.AID_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & signature_group == SIG_AID))$contribution)
age_60_plus.AID_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & signature_group == SIG_AID))$contribution)

age_45_under.not_AID_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & !(signature_group == SIG_AID)))$contribution)
age_45_to_60.not_AID_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & !(signature_group == SIG_AID)))$contribution)
age_60_plus.not_AID_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & !(signature_group == SIG_AID)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,3))
sam[1, ] <- c(age_45_under.AID_sig.num, age_45_to_60.AID_sig.num, age_60_plus.AID_sig.num) # 161, 21, 102
sam[2, ] <- c(age_45_under.not_AID_sig.num, age_45_to_60.not_AID_sig.num, age_60_plus.not_AID_sig.num) # 2649, 2437, 705

contingency_table[1:2,7:9] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 1.524559e-44
```

## Test association between age groups and the DNA mismatch repair signature

```{r}
age_45_under.DDMR_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & signature_group == SIG_DDMR))$contribution)
age_45_to_60.DDMR_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & signature_group == SIG_DDMR))$contribution)
age_60_plus.DDMR_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & signature_group == SIG_DDMR))$contribution)

age_45_under.not_DDMR_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & !(signature_group == SIG_DDMR)))$contribution)
age_45_to_60.not_DDMR_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & !(signature_group == SIG_DDMR)))$contribution)
age_60_plus.not_DDMR_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & !(signature_group == SIG_DDMR)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,3))
sam[1, ] <- c(age_45_under.DDMR_sig.num, age_45_to_60.DDMR_sig.num, age_60_plus.DDMR_sig.num) # 789, 643, 218
sam[2, ] <- c(age_45_under.not_DDMR_sig.num, age_45_to_60.not_DDMR_sig.num, age_60_plus.not_DDMR_sig.num) # 2022, 1815, 589

contingency_table[3:4,7:9] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 1.524559e-44
```

## Test association between age groups and the hypermutation signature

```{r}
age_45_under.hyper_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & signature_group == SIG_HYPER))$contribution)
age_45_to_60.hyper_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & signature_group == SIG_HYPER))$contribution)
age_60_plus.hyper_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & signature_group == SIG_HYPER))$contribution)

age_45_under.not_hyper_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & !(signature_group == SIG_HYPER)))$contribution)
age_45_to_60.not_hyper_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & !(signature_group == SIG_HYPER)))$contribution)
age_60_plus.not_hyper_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & !(signature_group == SIG_HYPER)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,3))
sam[1, ] <- c(age_45_under.hyper_sig.num, age_45_to_60.hyper_sig.num, age_60_plus.hyper_sig.num) # 77, 88, 0
sam[2, ] <- c(age_45_under.not_hyper_sig.num, age_45_to_60.not_hyper_sig.num, age_60_plus.not_hyper_sig.num) # 2734, 2370, 807

contingency_table[5:6,7:9] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 4.230702e-07
```

## Test association between age groups and the aging signature

```{r}
age_45_under.aging_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & signature_group == SIG_AGING))$contribution)
age_45_to_60.aging_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & signature_group == SIG_AGING))$contribution)
age_60_plus.aging_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & signature_group == SIG_AGING))$contribution)

age_45_under.not_aging_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & !(signature_group == SIG_AGING)))$contribution)
age_45_to_60.not_aging_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & !(signature_group == SIG_AGING)))$contribution)
age_60_plus.not_aging_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & !(signature_group == SIG_AGING)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,3))
sam[1, ] <- c(age_45_under.aging_sig.num, age_45_to_60.aging_sig.num, age_60_plus.aging_sig.num) # 659, 965, 157
sam[2, ] <- c(age_45_under.not_aging_sig.num, age_45_to_60.not_aging_sig.num, age_60_plus.not_aging_sig.num) # 2152, 1493, 650

contingency_table[7:8,7:9] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 9.696108e-45
```

## Test association between age groups and all other COSMIC signatures

```{r}
age_45_under.others_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & signature_group == SIG_OTHERS))$contribution)
age_45_to_60.others_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & signature_group == SIG_OTHERS))$contribution)
age_60_plus.others_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & signature_group == SIG_OTHERS))$contribution)

age_45_under.not_others_sig.num <- sum((filter(plot_data, age_group == AGE_45_UNDER & !(signature_group == SIG_OTHERS)))$contribution)
age_45_to_60.not_others_sig.num <- sum((filter(plot_data, age_group == AGE_45_TO_60 & !(signature_group == SIG_OTHERS)))$contribution)
age_60_plus.not_others_sig.num <- sum((filter(plot_data, age_group == AGE_60_PLUS & !(signature_group == SIG_OTHERS)))$contribution)

# create contingency table for chi-square test
sam <- array(dim = c(2,3))
sam[1, ] <- c(age_45_under.others_sig.num, age_45_to_60.others_sig.num, age_60_plus.others_sig.num) # 1125, 741, 331
sam[2, ] <- c(age_45_under.not_others_sig.num, age_45_to_60.not_others_sig.num, age_60_plus.not_others_sig.num) # 1686, 1717, 476

contingency_table[9:10,7:9] <- sam

# apply chi-square test to assess if hypermutation signature is linked to EBV status
pval <- chisq.test(sam)$p.value # 7.966147e-15
```

## Output results

```{r}
write.table(contingency_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/contingency-cosmic_signatures_vs_pheno_groups-mutation_level.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```

## PART 2 - mutations in a given gene and mutations in a given signature

```{r}

# data with gene-level information on mutations
plot_data <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/gene_level_path_groups_3+_mutated/contributions-gene_level-normalized.grouped.annotated.with_pathways.txt", sep = "\t", header = TRUE)

signatures <- as.vector(unique(plot_data$signature_group))
genes <- as.vector(unique(plot_data$gene))

NUM_signatures <- length(signatures)
NUM_genes <- length(genes)

contingency_table <- array(dim = c(NUM_signatures*2, NUM_genes*2))
i <- 1
j <- 1

pval_table <- array(dim = c(NUM_signatures, NUM_genes))
k <- 1
l <- 1

# for each gene
for (gene_name in genes) {
  
  # for each signature
  for (signature_name in signatures){
    
    # get values for contingency table
    num.gene_and_sig <- sum((filter(plot_data, 
                                    gene == gene_name & signature_group == signature_name))$contribution) 
    num.not_gene_and_sig <- sum((filter(plot_data, 
                                    gene != gene_name & signature_group == signature_name))$contribution) 
    num.gene_and_not_sig <- sum((filter(plot_data, 
                                    gene == gene_name & signature_group != signature_name))$contribution) 
    num.not_gene_and_not_sig <- sum((filter(plot_data, 
                                    gene != gene_name & signature_group != signature_name))$contribution) 
    
    # chi-square test
    sam <- array(dim = c(2,2))
    sam[1, ] <- c(num.gene_and_sig, num.not_gene_and_sig)
    sam[2, ] <- c(num.gene_and_not_sig, num.not_gene_and_not_sig)
    
    curr_row.start <- i
    curr_row.end <- i+1
    curr_col.start <- j
    curr_col.end <- j+1
    
    contingency_table[curr_row.start:curr_row.end, curr_col.start:curr_col.end] <- sam
    i <- i + 2
    
    # Chi-Square test
    pvalue <- chisq.test(sam)$p.value
    pval_table[k, l] <- pvalue
    k <- k + 1
  }
  # continue iterating through genes
  j <- j + 2
  l <- l + 1
  # reset signatures
  i <- 1
  k <- 1
}
  
colnames(pval_table) <- genes
row.names(pval_table) <- signatures

not.genes <- paste("Not ", genes, sep = "")
contingency.col_names <- c(rbind(genes, not.genes))
not.signatures <- paste("Not ", signatures, sep = "")
contingency.row_names <- c(rbind(signatures, not.signatures))

colnames(contingency_table) <- contingency.col_names
row.names(contingency_table) <- contingency.row_names

write.table(contingency_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/gene_level_path_groups_3+_mutated/gene_mutations_vs_key_cosmic_signatures-contingency_table.txt", sep = "\t", quote = FALSE)

write.table(pval_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/gene_level_path_groups_3+_mutated/chisq-gene_mutations_vs_key_cosmic_signatures.txt", sep = "\t", quote = FALSE)
```

## #########################################################################
## Visualizations
## #########################################################################

## Boxplot of normalized contributions of AID/APOBEC mutations in EBV+ vs. EBV- patients 

```{r}
library(ggplot2)

norm.contrib <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/eber_groups-norm_contrib-AID_APOBEC.txt", sep = "\t", header = TRUE)

p <- ggplot(norm.contrib, aes(x=EBER_status, y=total_norm_contrib)) + 
  geom_boxplot() +
  ylab("Normalized contribution of mutations to AID / APOBEC signature") + 
  xlab("EBER status") +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.7)
  
#stat_summary(fun.y=mean, geom="point", shape=23, size=4)

ebv_pos <- filter(norm.contrib, EBER_status == "Positive")
ebv_neg <- filter(norm.contrib, EBER_status == "Negative")

# t-test to see if the distributions are significantly different
t.test(ebv_pos$total_norm_contrib, ebv_neg$total_norm_contrib)

#	Welch Two Sample t-test
#
#data:  ebv_pos$total_norm_contrib and ebv_neg$total_norm_contrib
#t = -2.3798, df = 27.617, p-value = 0.02447
#alternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -0.092384630 -0.006885817
#sample estimates:
# mean of x  mean of y 
#0.02208149 0.07171671 

```

