---
title: "MutationalPatterns for GZL exomes"
author: "Stacy Hung"
date: "April 19, 2019"
output: html_document
---

This script applies the MutationalPatterns package to understand mutational signatures and patterns of base nucleotide substitutions in the GZL exome cohort.

This code must be run from Rogue, and referencing files in /data/projects/stacy/gzl_exomes/signature_analysis/

## Load data

```{r}
library("BSgenome.Hsapiens.UCSC.hg19", character.only = TRUE)
library(MutationalPatterns) 
library(NMF)

ref_genome <- "BSgenome.Hsapiens.UCSC.hg19"

# locate VCF files
vcf_files <- list.files(path = "/data/projects/stacy/gzl_exomes/signature_analysis/input", pattern = "*.vcf", full.names = TRUE, all.files = TRUE)
# read in sample names and metadata to annotate the VCF files (remove GZ310 - sample id "GZ310T")
metadata <- read.table("/data/projects/stacy/gzl_exomes/signature_analysis/data/metadata.txt", sep = "\t", header = TRUE)
sample_names <- metadata$tumor_id
# load VCF files
vcf <- read_vcfs_as_granges(vcf_files = vcf_files, sample_names = sample_names, ref_genome)
```

## Mutational characteristics

```{r}

## Base substitution types

# retrieve base substitutions from the VCF GRanges object as REF>ALT
muts <- mutations_from_vcf(vcf[[1]])
# Convert base substitutions to the 6 types of conventional base types
types <- mut_type(vcf[[1]])
# retrieve the sequence context (one base upstream and one base downstream)
context <- mut_context(vcf[[1]], ref_genome)
# retrieve the type and context for all positions in the VCF GRanges object
type_context <- type_context(vcf[[1]], ref_genome)
# count mutation type occurrences for all VCF objects in the GRangesList
type_occurrences <- mut_type_occurrences(vcf, ref_genome)

## Mutation Spectrum

# plot the mean relative contribution of each of the 6 base substitution types over all sampels
p1 <- plot_spectrum(type_occurrences)
# plot mutation spectrum with distinction between C>T at CpG sites and other sites
p2 <- plot_spectrum(type_occurrences, CT=TRUE)
# combine the plots
library(gridExtra)
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/mutation_spectrum.pdf", width=6, height = 8) 
grid.arrange(p1, p2, ncol=1, widths=c(4,4), heights=c(2,2))
dev.off() 
# facet by sample group - EBV
p4 <- plot_spectrum(type_occurrences, by = metadata$EBV_status, CT = TRUE, legend = TRUE)
# facet by sample group - path group
p5 <- plot_spectrum(type_occurrences, by = metadata$path_group, CT = TRUE, legend = TRUE)
# facet by sample group - mediastinal involvement
p6 <- plot_spectrum(type_occurrences, by = metadata$mediastinal_involvement, CT = TRUE, legend = TRUE)
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/mutation_spectrum-facets.pdf", height = 14) 
grid.arrange(p4, p5, p6, ncol=1, heights=c(3, 6, 3))
dev.off() 

### 96 mutational profile

# generate a 96-trinucleotide mutation count matrix
mut_mat <- mut_matrix(vcf_list = vcf, ref_genome = ref_genome)
# plot the 96 profile of all the samples
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/plot_96_profile-30_samples.pdf", height = 24) 
plot_96_profile(mut_mat)
dev.off() 

```

## Mutational signatures

Mutational signatures can be extracted from the mutation count matrix with NMF.  A critical parameter in NMF is the factorization rank, which is the number of mutational signatures.  You can find the optimal factorization rank using the NMF package. The most common approach is to choose the smallest rank for which the cophenetic correlation coefficient starts decreasing.

```{r}

## De novo mutational signature extraction using NMF


# first add a small pseudocount to mutation count matrix
mut_mat <- mut_mat + 0.0001
# use the NMF package to generate an estimate rank plot
estimate <- nmf(mut_mat, rank = 1:10, method = "brunet", nrun = 100, seed = 123456) # doesn't work on Rogue
# plot the estimates
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/plot_estimate.pdf") 
plot(estimate); dev.off() 


# optimal rank appears to be 2
# perform a relative large number iterations to achieve stability and avoid local minima
# recommended number of iterations (runs) is 400-500 based on algorithm paper by Alexandrov et al.
nmf_res <- extract_signatures(mut_mat, rank = 2, nrun = 400)
# assign signature names
colnames(nmf_res$signatures) <- c("GZ Signature A", "GZ Signature B")
rownames(nmf_res$contribution) <- c("GZ Signature A", "GZ Signature B")
# print out contributions and signatures to output files
write.table(nmf_res$contributions, "/data/projects/stacy/gzl_exomes/signature_analysis/output/28_samples_analysis/denovo-contributions.txt", sep = "\t", quote = FALSE)
write.table(nmf_res$signature, "/data/projects/stacy/gzl_exomes/signature_analysis/output/28_samples_analysis/denovo-signatures.txt", sep = "\t", quote = FALSE)

# plot the 96-profile of the signatures
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/28_samples/de_novo-plot_96_profile.pdf", height = 3.5) 
plot_96_profile(nmf_res$signatures); dev.off() 

# visualize the contribution of the signatures in a barplot
# order by specific groups of interest (contribution is a 2X28 matrix where 2 = signatures; 28 = samples)
col.order <- c("GZ-095", "GZ-BCC-020", "GZ-178", "GZ-048", "GZ-180", "GZ-BCC-013", "GZ-116", "GZ-099", "GZ-046", "GZ-BCC-054", "GZ-176", "GZ-184", "GZ-152", "GZ-044", "GZ-149", "GZ-092", "GZ-230", "GZ-086", "GZ-267", "GZ-068", "GZ-062", "GZ-197", "GZ-BCC-008", "GZ-235", "GZ-064", "GZ-294", "GZ-301", "GZ-032")
nmf_res$contribution <- nmf_res$contribution[, col.order]

# visualize the contribution in absolute number of mutations
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/28_samples/signature-absolute_contributions.pdf", width = 5, height = 5) 
plot_contribution(nmf_res$contribution, nmf_res$signature, mode = "absolute", coord_flip=TRUE); dev.off()
# visualize the contribution in relative number of mutations
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/28_samples/signature-relative_contributions.pdf", width = 5, height = 5) 
plot_contribution(nmf_res$contribution, nmf_res$signature, mode = "relative", coord_flip=TRUE); dev.off()


# Plot signature contribution as a heatmap with sample clustering dendogram and specified signature order:
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/28_samples/signature-heatmap_contributions.pdf", width = 3, height = 8) 
plot_contribution_heatmap(nmf_res$contribution, sig_order = c("GZ Signature A", "GZ Signature B"), cluster_samples=FALSE, plot_values=TRUE); dev.off()
# calculate the relative contributions
denovo.contributions <- as.matrix(read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/28_samples_analysis/denovo-contributions.txt", sep = "\t", header = TRUE))
# transpose
denovo.contributions <- t(denovo.contributions)
# relative contribution
denovo.contrib.norm <- denovo.contributions / rowSums(denovo.contributions)
write.table(denovo.contrib.norm, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/28_samples_analysis/denovo-contributions.norm.txt", sep = "\t", quote = FALSE)

# compare reconstructed mutational profile with original mutational profile
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/28_samples/comparison_of_profiles.pdf") 
plot_compare_profiles(mut_mat[,1], nmf_res$reconstructed[,1], profile_names = c("Original", "Reconstructed"), condensed = TRUE); dev.off()


## COSMIC mutational signatures


# Download mutational signatures from the COSMIC website (saved local directory to avoid connection issues)
cancer_signatures <- read.table("/data/projects/stacy/gzl_exomes/signature_analysis/signatures_probabilities.txt", sep = "\t", header = TRUE)
# Match the order of the mutation types to MutationalPatterns standard
new_order = match(row.names(mut_mat), cancer_signatures$Somatic.Mutation.Type)
# Reorder cancer signatures dataframe
cancer_signatures = cancer_signatures[as.vector(new_order),]
# Add trinucletiode changes names as row.names
row.names(cancer_signatures) = cancer_signatures$Somatic.Mutation.Type
# Keep only 96 contributions of the 30 signatures in matrix
cancer_signatures = as.matrix(cancer_signatures[,4:33])
# plot mutational profile of the COSMIC signatures
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/COSMIC_signatures.pdf") 
plot_96_profile(cancer_signatures, condensed = TRUE, ymax = 0.3); dev.off()
# hierarchically cluster the COSMIC signatures based on their similarity with average linkage:
hcluster_cosmic <- cluster_signatures(cancer_signatures, method = "average")
cosmic_order <- colnames(cancer_signatures)[hcluster_cosmic$order]
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/hclust-COSMIC.pdf") 
plot(hcluster_cosmic); dev.off()


## Similarity between de novo signatures and COSMIC signatures

# calculate pairwise cosine similarity between mutational profiles and COSMIC signatures
cos_sim_denovo_signatures <- cos_sim_matrix(nmf_res$signatures, cancer_signatures)
# plot heatmap of cosine similarities
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/heatmap-COSMIC_vs_denovo_signatures.pdf", width = 8, height = 2) 
plot_cosine_heatmap (cos_sim_denovo_signatures, col_order = cosmic_order, cluster_rows = FALSE, plot_values=TRUE); dev.off()

## plot signature contribution as heatmap
pheatmap1 <- plot_contribution_heatmap(nmf_res$contribution, cluster_samples=FALSE)

## Similarity between mutational profiles and COSMIC signatures

# calculate pairwise cosine similarity between mutational profiles and COSMIC signatures
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat, cancer_signatures)
# plot heatmap with specific signature order
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/heatmap-COSMIC_vs_GZ_signatures.pdf", width = 12, height = 7) 
plot_cosine_heatmap (cos_sim_samples_signatures, col_order = cosmic_order, cluster_rows = TRUE); dev.off()

# Find optimal contribution of COSMIC signatures to reconstruct 96 mutational profiles

# fit mutation matrix to the COSMIC mutational signatures
fit_res <- fit_to_signatures(mut_mat, cancer_signatures)
# print contributions to output
write.table(fit_res$contributions, "/data/projects/stacy/gzl_exomes/signature_analysis/output/30_samples_analysis/contributions.txt", sep = "\t", quote = FALSE)
# print reconstructed to output
write.table(fit_res$reconstructed, "/data/projects/stacy/gzl_exomes/signature_analysis/output/30_samples_analysis/reconstructed.txt", sep = "\t", quote = FALSE)

# plot optimal contribution of the COSMIC signatures in each sample as stacked barplot
# first select signatures with a minimum of X mutations (summed across all samples) contributing to that signature
select <- which (rowSums(fit_res$contribution) >= 300)
# plot cases by EBV status
contributions.sorted <- as.matrix(read.table("/data/projects/stacy/gzl_exomes/signature_analysis/input/signatures_of_interest/optimal-contributions-sorted_by_EBER_and_num_mutations.txt", sep="\t",header=TRUE, row.names = 1))
# plot contribution barplot
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/COSMIC_contributions-per_sample.sorted.pdf", width = 12, height = 7)
#plot_contribution(fit_res$contribution[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "absolute"); dev.off()
plot_contribution(contributions.sorted[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "absolute"); dev.off()
# and the same plot, but for relative contributions
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/COSMIC_relative_contributions-per_sample.sorted.pdf", width = 12, height = 7)
plot_contribution(contributions.sorted[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "relative"); dev.off()


# write out the results so that we can select exactly what we want to plot
temp <- fit_res$contribution
temp2 <- cancer_signatures
write.table(temp, "/data/projects/stacy/gzl_exomes/signature_analysis/output/30_samples_analysis/GZ-optimal_contributions-COSMIC.txt", sep = "\t", quote = FALSE)
write.table(temp2, "/data/projects/stacy/gzl_exomes/signature_analysis/output/30_samples_analysis/COSMIC-signatures.txt", sep = "\t", quote = FALSE)

# same plot, but with patients sorted by EBV status (or other grouping)
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/COSMIC_contributions-per_sample-sorted_by_EBV.pdf", width = 12, height = 7)
plot_contribution(fit_res$contribution[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "absolute")
dev.off()

# same plot with patients sorted by EBV status and COSMIC signatures of interest (read in manually)
#signatures.of.interest <- read.table("/data/projects/stacy/gzl_exomes/signature_analysis/input/signatures_of_interest/signatures_of_interest.txt", sep="\t",header=TRUE)
signatures.of.interest <- read.table("/data/projects/stacy/gzl_exomes/signature_analysis/input/signatures_of_interest/signatures_of_interest.sorted.txt", sep="\t",header=TRUE)
select <- signatures.of.interest$signature_index
names(select) <- signatures.of.interest$signature_name
# read in optimal contributions that have been sorted by EBV status
contributions.sorted <- as.matrix(read.table("/data/projects/stacy/gzl_exomes/signature_analysis/input/signatures_of_interest/optimal-contributions-sorted_by_EBER_and_num_mutations.txt", sep="\t",header=TRUE, row.names = 1))
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/COSMIC_contributions-per_sample-sorted_by_EBV-signatures_of_interest.pdf", width = 12, height = 7)
plot_contribution(contributions.sorted[select,], 
                  cancer_signatures[,select], 
                  coord_flip = TRUE, 
                  mode = "absolute", 
                  palette = c("lightsalmon", "lightsalmon2", "mediumturquoise", "mediumpurple1", "mediumpurple2", "mediumpurple3", "mediumpurple4")); dev.off()
# and relative contribution plot
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/COSMIC_relative_contributions-per_sample-sorted_by_EBV-signatures_of_interest.pdf", width = 12, height = 7)
plot_contribution(contributions.sorted[select,], 
                  cancer_signatures[,select], 
                  coord_flip = TRUE, 
                  mode = "relative", 
                  palette = c("lightsalmon", "lightsalmon2", "mediumturquoise", "mediumpurple1", "mediumpurple2", "mediumpurple3", "mediumpurple4")); dev.off()

# plot same plot but without GZ310
# first remove GZ310 from the mutation matrix
mut_mat_minus_310 <- mut_mat[, -(27)]
fit_res_minus_310 <- fit_to_signatures(mut_mat_minus_310, cancer_signatures)
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/COSMIC_contributions-per_sample_minus_310.pdf", width = 12, height = 7)
plot_contribution(fit_res_minus_310$contribution[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "absolute")
dev.off()

# plot relative contribution barplot
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/COSMIC_optimal_relative_contributions-per_sample.pdf", width = 12, height = 7)
plot_contribution(fit_res$contribution[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "relative")
dev.off()

# plot relative contribution of cancer signatures in each sample as a heatmap with sample clustering
pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/30_samples/heatmap-COSMIC_contributions-per_sample.pdf", width = 10, height = 6)
plot_contribution_heatmap(fit_res$contribution, cluster_samples = TRUE, method = "complete")
dev.off()
```


# Establish distribution of contributions across all patients and signatures

```{r}
library(tidyr)

optimal_contributions <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/GZ-optimal_contributions-COSMIC.txt", sep = "\t", header = TRUE, row.names = 1, stringsAsFactors = FALSE)

optimal_contributions.matrix <- as.data.frame(t(optimal_contributions))
mutation_sums <- rowSums(optimal_contributions.matrix)

# normalized contributions
optimal_contributions.matrix.norm <- optimal_contributions.matrix/mutation_sums

# print out to file
write.table(optimal_contributions.matrix.norm, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/GZ-optimal_contributions-COSMIC.normalized.txt", sep = "\t", quote = FALSE)

# convert from wide to long
#optimal_contributions.long <- gather(optimal_contributions, sample, contribution, GZ.176:GZ.086)
optimal_contributions.long <- gather(optimal_contributions.matrix.norm, signature, contribution, Signature.1:Signature.30)

# order by contributions
optimal_contributions.long <- optimal_contributions.long[order(optimal_contributions.long$contribution), ]

# plot density of contributions
p1 <- ggplot(optimal_contributions.long, aes(as.numeric(contribution))) + 
  geom_density() +
  xlab("Normalized contribution") + 
  ylab("Density") +
  geom_vline(xintercept = 0.02, col="red", lty=2) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
  ggtitle("All mutations (n=900 observations)")

p2 <- ggplot(optimal_contributions.long, aes(as.numeric(contribution))) + 
  geom_density() +
  xlab("Mutation contribution") + 
  ylab("Density") +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank()) +
  xlim(0.02, 0.4) + 
  geom_vline(xintercept = 0.05, col="blue", lty=2) +
  ggtitle("0.02 < Mutations with norm contribution < 0.4 (per sample/signature) (n=265 observations)")
  
grid.arrange(p1, p2, ncol=1)

```

## Create new contribution matrix based on contribution cutoff and grouping of top and key signatures

Categories: 
(1) Aging - #1
(2) AID/APOBEC activity - #2, 13
(3) Defective DNA mismatch repair - #3, 6, 15, 20, 26
(4) Polymerase-n-mediated somatic hypermutation - #9
(5) All other signatures

Patients will be ordered first by EBV group, then by increasing numbers of mutations - all mutations should be captured by the categories listed above.

```{r}
library(ggplot2)

optimal_contributions <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/GZ-optimal_contributions-COSMIC.txt", sep = "\t", header = TRUE, row.names = 1, stringsAsFactors = FALSE)

# create new contribution matrix with just the 5 signature groups
df <- as.data.frame(t(optimal_contributions))

contributions.group1 <- df$Signature.1
contributions.group2 <- df$Signature.2 + df$Signature.13
contributions.group3 <- df$Signature.3 + df$Signature.6 + df$Signature.15 + df$Signature.20 + df$Signature.26
contributions.group4 <- df$Signature.9
contributions.group5 <- df$Signature.4 + df$Signature.5 + df$Signature.7 + df$Signature.8 + df$Signature.10 + df$Signature.11 + df$Signature.12 + df$Signature.14 + df$Signature.16 + df$Signature.17 + df$Signature.18 + df$Signature.19 + df$Signature.21 + df$Signature.22 + df$Signature.23 + df$Signature.24 + df$Signature.25 + df$Signature.27 + df$Signature.28 + df$Signature.29 + df$Signature.30

contributions.grouped <- as.data.frame(cbind(row.names(df), contributions.group1, contributions.group2, contributions.group3, contributions.group4, contributions.group5))

contributions.grouped.without_names <- as.data.frame(cbind(contributions.group1, contributions.group2, contributions.group3, contributions.group4, contributions.group5))
total_mutations.by_group <- colSums(contributions.grouped.without_names)

# to verify that all the signatures have been included
#total_mutations <- rowSums(contributions.grouped)

# convert from wide to long
contributions.grouped.long <- gather(contributions.grouped, 
                                     signature_group, 
                                     contribution, 
                                     contributions.group1:contributions.group5)
colnames(contributions.grouped.long) <- c("patient", "signature_group", "contribution")

write.table(contributions.grouped.long, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/GZ-optimal_contributions-COSMIC.normalized.grouped.txt", sep = "\t", quote = FALSE)

plot_data <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/GZ-optimal_contributions-COSMIC.normalized.grouped.eber_status.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# specify order of patients:
patients.order <- c("GZ-048", "GZ-180", "GZ-178", "GZ-BCC-020", "GZ-116", "GZ-BCC-013", "GZ-BCC-054", "GZ-176", "GZ-046", "GZ-184", "GZ-149", "GZ-044", "GZ-230", "GZ-092", "GZ-152", "GZ-095", "GZ-267", "GZ-086", "GZ-099", "GZ-068", "GZ-222", "GZ-062", "GZ-310", "", "", "GZ-197", "GZ-BCC-008", "GZ-235", "GZ-294", "GZ-064", "GZ-301", "GZ-032")
# specify order of signature groups:
sig_groups.order <- c("Defective DNA mismatch repair (3, 6, 15, 20, 26)",
                      "AID / APOBEC activity (2, 13)",
                      "Polymerase-n-mediated somatic hypermutation (9)", 
                      "Aging (1)",
                      "All other COSMIC signatures")

p1 <- ggplot(data = plot_data, 
             aes(factor(x = patient, levels = patients.order), 
                 y = as.numeric(contribution), 
                 fill = factor(signature_group, levels = sig_groups.order),
                 color = EBER_status)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(name = "Signature Group",
                    values = c("Aging (1)" = "rosybrown1",
                               "AID / APOBEC activity (2, 13)" = "lightgreen", 
                               "Defective DNA mismatch repair (3, 6, 15, 20, 26)" = "mediumpurple1",
                               "Polymerase-n-mediated somatic hypermutation (9)" = "dodgerblue", 
                               "All other COSMIC signatures" = "gray")) +
  scale_colour_manual(name = "EBER status", 
                      values = c("Negative"="black", "Positive"="red")) +
  xlab("") + 
  ylab("Number of mutations")
  
# for a very fine outline (use inside geom_bar):
# size = 0.2

signature_groups.order <- c("All other COSMIC signatures", 
                            "Aging (1)", 
                            "Defective DNA mismatch repair (3, 6, 15, 20, 26)",
                            "AID / APOBEC activity (2, 13)",
                            "Polymerase-n-mediated somatic hypermutation (9)")

# barplot showing total number of mutations by signature (and stacked by EBV- vs. EBV+ cases)
p2 <- ggplot(data = plot_data, 
             aes(factor(x = signature_group, levels = rev(signature_groups.order)),
                 y = as.numeric(contribution), 
                 fill = EBER_status)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(name = "EBER status",
                    values = c("Negative" = "black",
                               "Positive" = "red")) +
  xlab("") + 
  ylab("Number of mutations") +
  coord_flip()

```


## Calculate enrichment of the AID/APOBEC signature in EBV vs. EBV- groups

```{r}
library(dplyr)

EBV_pos.AID_sig.data <- filter(plot_data, EBER_status == "Positive" & signature_group == "AID / APOBEC activity (2, 13)")
EBV_pos.not_AID_sig.data <- filter(plot_data, EBER_status == "Positive" & !(signature_group == "AID / APOBEC activity (2, 13)"))
EBV_neg.AID_sig.data <- filter(plot_data, EBER_status == "Negative" & signature_group == "AID / APOBEC activity (2, 13)")
EBV_neg.not_AID_sig.data <- filter(plot_data, EBER_status == "Negative" & !(signature_group == "AID / APOBEC activity (2, 13)"))

EBV_pos.AID_sig.num <- sum(EBV_pos.AID_sig.data$contribution) # 17
EBV_pos.not_AID_sig.num <- sum(EBV_pos.not_AID_sig.data$contribution) # 633
EBV_neg.AID_sig.num <- sum(EBV_neg.AID_sig.data$contribution) # 267
EBV_neg.not_AID_sig.num <- sum(EBV_neg.not_AID_sig.data$contribution) # 5158

EBV_pos.total <- EBV_pos.AID_sig.num + EBV_pos.not_AID_sig.num
EBV_neg.total <- EBV_neg.AID_sig.num + EBV_neg.not_AID_sig.num

# chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.AID_sig.num, EBV_neg.AID_sig.num)
sam[2, ] <- c(EBV_pos.total, EBV_neg.total)
  
# Chi-Square test to assess if AID/APOBEC in EBV+ cases is more different than expected
pvalue <- chisq.test(sam)$p.value # 0.01326989

```

## Boxplot of normalized contributions of AID/APOBEC mutations in EBV+ vs. EBV- patients 

```{r}
library(ggplot2)

norm.contrib <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/eber_groups-norm_contrib-AID_APOBEC.txt", sep = "\t", header = TRUE)

p <- ggplot(norm.contrib, aes(x=EBER_status, y=total_norm_contrib)) + 
  geom_boxplot() +
  ylab("Normalized contribution of mutations to AID / APOBEC signature") + 
  xlab("EBER status") +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.7)
  
#stat_summary(fun.y=mean, geom="point", shape=23, size=4)

ebv_pos <- filter(norm.contrib, EBER_status == "Positive")
ebv_neg <- filter(norm.contrib, EBER_status == "Negative")

# t-test to see if the distributions are significantly different
t.test(ebv_pos$total_norm_contrib, ebv_neg$total_norm_contrib)

#	Welch Two Sample t-test
#
#data:  ebv_pos$total_norm_contrib and ebv_neg$total_norm_contrib
#t = -2.3798, df = 27.617, p-value = 0.02447
#alternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -0.092384630 -0.006885817
#sample estimates:
# mean of x  mean of y 
#0.02208149 0.07171671 

```

## Barplot of cosine similarities between original and reconstructed profile - i.e. show how well each mutational profile can be reconstructed with the COSMIC mutational signatures (identical profiles = 1) - e.g. use 0.95 as a cutoff

```{r}
library(ggplot2)

# first calculate all pairwise cosine similarities
cos_sim_ori_rec <- cos_sim_matrix(mut_mat, fit_res$reconstructed)
# extract cosine similarities per sample between original and reconstructed
cos_sim_ori_rec <- as.data.frame(diag(cos_sim_ori_rec))
# adjust data frame for plotting with ggplot
colnames(cos_sim_ori_rec) = "cos_sim"
cos_sim_ori_rec$sample <- row.names(cos_sim_ori_rec)
# make barplot

# sort by decreasing similarity
#cos_sim_ori_rec <- within(cos_sim_ori_rec, sample <- factor(sample, levels=names(sort(cos_sim), decreasing=TRUE)))

pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/28_samples/cosmic_vs_reconstructed-cosine_similarity_barplot.pdf", width = 8, height = 7)

## plot
ggplot(cos_sim_ori_rec, aes(y=cos_sim, x=reorder(sample, -cos_sim))) + 
  geom_bar(stat="identity", fill="skyblue4") + 
  coord_flip(ylim=c(0.7, 1)) +
  ylab("Cosine similarity\n original VS reconstructed") +
  xlab("") +
  theme_bw() +
  theme(panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank()) +
  geom_hline(aes(yintercept=0.80), linetype = "dashed")

dev.off()

```

