---
title: "MutationalPatterns for GZL exomes"
author: "Stacy Hung"
date: "April 19, 2019"
output: html_document
---

This script analyzes mutations associated with a given phenotype and COSMIC signature (e.g. what is the signficance of mutations attributable to AID/APOBEC activity and EBV status?).

COSMIC signature categories: 
(1) Aging - #1
(2) AID/APOBEC activity - #2, 13
(3) Defective DNA mismatch repair - #3, 6, 15, 20, 26
(4) Polymerase-n-mediated somatic hypermutation - #9
(5) All other signatures

Phenotypic groups:
(1) EBV status (+/-)
(2) Mediastinal status (+/-)
(3) Path group (1/2 or 0/3)
(3) Age group (<= 45, 45-60, >= 60)


## Analyze associations 

```{r}
library(ggplot2)
library(tidyr)

plot_data <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/GZ-optimal_contributions-COSMIC.normalized.grouped.eber_status.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# specify order of patients:
patients.order <- c("GZ-048", "GZ-180", "GZ-178", "GZ-BCC-020", "GZ-116", "GZ-BCC-013", "GZ-BCC-054", "GZ-176", "GZ-046", "GZ-184", "GZ-149", "GZ-044", "GZ-230", "GZ-092", "GZ-152", "GZ-095", "GZ-267", "GZ-086", "GZ-099", "GZ-068", "GZ-222", "GZ-062", "GZ-310", "GZ-197", "GZ-BCC-008", "GZ-235", "GZ-294", "GZ-064", "GZ-301", "GZ-032")
# specify order of signature groups:
sig_groups.order <- c("Defective DNA mismatch repair (3, 6, 15, 20, 26)",
                      "AID / APOBEC activity (2, 13)",
                      "Polymerase-n-mediated somatic hypermutation (9)", 
                      "Aging (1)",
                      "All other COSMIC signatures")

p1 <- ggplot(data = plot_data, 
             aes(factor(x = patient, levels = patients.order), 
                 y = as.numeric(contribution), 
                 fill = factor(signature_group, levels = sig_groups.order),
                 color = EBER_status)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(name = "Signature Group",
                    values = c("Aging (1)" = "rosybrown1",
                               "AID / APOBEC activity (2, 13)" = "lightgreen", 
                               "Defective DNA mismatch repair (3, 6, 15, 20, 26)" = "mediumpurple1",
                               "Polymerase-n-mediated somatic hypermutation (9)" = "dodgerblue", 
                               "All other COSMIC signatures" = "gray")) +
  scale_colour_manual(name = "EBER status", 
                      values = c("Negative"="black", "Positive"="red")) +
  xlab("") + 
  ylab("Number of mutations")
  
# for a very fine outline (use inside geom_bar):
# size = 0.2

signature_groups.order <- c("All other COSMIC signatures", 
                            "Aging (1)", 
                            "Defective DNA mismatch repair (3, 6, 15, 20, 26)",
                            "AID / APOBEC activity (2, 13)",
                            "Polymerase-n-mediated somatic hypermutation (9)")

# barplot showing total number of mutations by signature (and stacked by EBV- vs. EBV+ cases)
p2 <- ggplot(data = plot_data, 
             aes(factor(x = signature_group, levels = rev(signature_groups.order)),
                 y = as.numeric(contribution), 
                 fill = EBER_status)) + 
  geom_bar(stat = "identity", width = 0.6) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(name = "EBER status",
                    values = c("Negative" = "black",
                               "Positive" = "red")) +
  xlab("") + 
  ylab("Number of mutations") +
  coord_flip()

```


## Calculate enrichment of the AID/APOBEC signature in EBV vs. EBV- groups

```{r}
library(dplyr)

EBV_pos.AID_sig.data <- filter(plot_data, EBER_status == "Positive" & signature_group == "AID / APOBEC activity (2, 13)")
EBV_pos.not_AID_sig.data <- filter(plot_data, EBER_status == "Positive" & !(signature_group == "AID / APOBEC activity (2, 13)"))
EBV_neg.AID_sig.data <- filter(plot_data, EBER_status == "Negative" & signature_group == "AID / APOBEC activity (2, 13)")
EBV_neg.not_AID_sig.data <- filter(plot_data, EBER_status == "Negative" & !(signature_group == "AID / APOBEC activity (2, 13)"))

EBV_pos.AID_sig.num <- sum(EBV_pos.AID_sig.data$contribution) # 17
EBV_pos.not_AID_sig.num <- sum(EBV_pos.not_AID_sig.data$contribution) # 633
EBV_neg.AID_sig.num <- sum(EBV_neg.AID_sig.data$contribution) # 267
EBV_neg.not_AID_sig.num <- sum(EBV_neg.not_AID_sig.data$contribution) # 5158

EBV_pos.total <- EBV_pos.AID_sig.num + EBV_pos.not_AID_sig.num
EBV_neg.total <- EBV_neg.AID_sig.num + EBV_neg.not_AID_sig.num

# chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.AID_sig.num, EBV_neg.AID_sig.num)
sam[2, ] <- c(EBV_pos.total, EBV_neg.total)
  
# Chi-Square test to assess if AID/APOBEC in EBV+ cases is more different than expected
pvalue <- chisq.test(sam)$p.value # 0.01326989

```

## Calculate enrichment of defective DNA mismatch repair signature in EBV+ vs. EBV- groups

```{r}
library(dplyr)

EBV_pos.DDMR_sig.data <- filter(plot_data, EBER_status == "Positive" & signature_group == "Defective DNA mismatch repair (3, 6, 15, 20, 26)")
EBV_pos.not_DDMR_sig.data <- filter(plot_data, EBER_status == "Positive" & !(signature_group == "Defective DNA mismatch repair (3, 6, 15, 20, 26)"))
EBV_neg.DDMR_sig.data <- filter(plot_data, EBER_status == "Negative" & signature_group == "Defective DNA mismatch repair (3, 6, 15, 20, 26)")
EBV_neg.not_DDMR_sig.data <- filter(plot_data, EBER_status == "Negative" & !(signature_group == "Defective DNA mismatch repair (3, 6, 15, 20, 26)"))

EBV_pos.DDMR_sig.num <- sum(EBV_pos.DDMR_sig.data$contribution) # 191.3075
EBV_pos.not_DDMR_sig.num <- sum(EBV_pos.not_DDMR_sig.data$contribution) # 458.9504
EBV_neg.DDMR_sig.num <- sum(EBV_neg.DDMR_sig.data$contribution) # 1458.566
EBV_neg.not_DDMR_sig.num <- sum(EBV_neg.not_DDMR_sig.data$contribution) # 3966.872

EBV_pos.total <- EBV_pos.DDMR_sig.num + EBV_pos.not_DDMR_sig.num
EBV_neg.total <- EBV_neg.DDMR_sig.num + EBV_neg.not_DDMR_sig.num

# chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.DDMR_sig.num, EBV_neg.DDMR_sig.num)
sam[2, ] <- c(EBV_pos.total, EBV_neg.total)
  
# Chi-Square test to assess if DDMR in EBV+ cases is more different than expected
pvalue <- chisq.test(sam)$p.value # 0.3233782


```

## Boxplot of normalized contributions of AID/APOBEC mutations in EBV+ vs. EBV- patients 

```{r}
library(ggplot2)

norm.contrib <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/eber_groups-norm_contrib-AID_APOBEC.txt", sep = "\t", header = TRUE)

p <- ggplot(norm.contrib, aes(x=EBER_status, y=total_norm_contrib)) + 
  geom_boxplot() +
  ylab("Normalized contribution of mutations to AID / APOBEC signature") + 
  xlab("EBER status") +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.7)
  
#stat_summary(fun.y=mean, geom="point", shape=23, size=4)

ebv_pos <- filter(norm.contrib, EBER_status == "Positive")
ebv_neg <- filter(norm.contrib, EBER_status == "Negative")

# t-test to see if the distributions are significantly different
t.test(ebv_pos$total_norm_contrib, ebv_neg$total_norm_contrib)

#	Welch Two Sample t-test
#
#data:  ebv_pos$total_norm_contrib and ebv_neg$total_norm_contrib
#t = -2.3798, df = 27.617, p-value = 0.02447
#alternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -0.092384630 -0.006885817
#sample estimates:
# mean of x  mean of y 
#0.02208149 0.07171671 

```

## Barplot of cosine similarities between original and reconstructed profile - i.e. show how well each mutational profile can be reconstructed with the COSMIC mutational signatures (identical profiles = 1) - e.g. use 0.95 as a cutoff

```{r}
library(ggplot2)

# first calculate all pairwise cosine similarities
cos_sim_ori_rec <- cos_sim_matrix(mut_mat, fit_res$reconstructed)
# extract cosine similarities per sample between original and reconstructed
cos_sim_ori_rec <- as.data.frame(diag(cos_sim_ori_rec))
# adjust data frame for plotting with ggplot
colnames(cos_sim_ori_rec) = "cos_sim"
cos_sim_ori_rec$sample <- row.names(cos_sim_ori_rec)
# make barplot

# sort by decreasing similarity
#cos_sim_ori_rec <- within(cos_sim_ori_rec, sample <- factor(sample, levels=names(sort(cos_sim), decreasing=TRUE)))

pdf("/data/projects/stacy/gzl_exomes/signature_analysis/figures/all_samples/28_samples/cosmic_vs_reconstructed-cosine_similarity_barplot.pdf", width = 8, height = 7)

## plot
ggplot(cos_sim_ori_rec, aes(y=cos_sim, x=reorder(sample, -cos_sim))) + 
  geom_bar(stat="identity", fill="skyblue4") + 
  coord_flip(ylim=c(0.7, 1)) +
  ylab("Cosine similarity\n original VS reconstructed") +
  xlab("") +
  theme_bw() +
  theme(panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank()) +
  geom_hline(aes(yintercept=0.80), linetype = "dashed")

dev.off()

```

