---
title: "MutationalPatterns for GZL exomes"
author: "Stacy Hung"
date: "April 19, 2019"
output: html_document
---

This script applies the MutationalPatterns package to understand mutational signatures and patterns of base nucleotide substitutions in the GZL exome cohort.

## Preprocessing steps:

## 1. Modify master MAF file to be used to generate VCF files so that tumor names are actually gene names

```{r}
library(dplyr)

maf <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/input/snvs_indels.full_cohort_minus_GZ229.default_and_optimized.incl_silent_and_UTR.for_vcf_conversion.maf", sep = "\t", header = TRUE, fill = TRUE)

# reorder columns
keeps <- c("Tumor_Sample_Barcode", "Hugo_Symbol", "Variant_Classification", "Variant_Type", "Reference_Allele",   "Chromosome", "Start_Position", "Tumor_Seq_Allele1", "Tumor_Seq_Allele2", "End_Position")
maf <- maf[keeps]

# Exclude GZ310 mutations
maf <- filter(maf, maf$Tumor_Sample_Barcode != "GZ-310")

# remove indels since they will not be analyzed
maf <- filter(maf, maf$Variant_Type == "SNP")

MIN_CONTRIBUTION = nrow(maf)*0.03; # "top"-contributing signatures based on contributing to >= 3% of all mutations

write.table(maf, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/input/snvs.full_cohort_minus_GZ229_and_GZ310.default_and_optimized.incl_silent_and_UTR.for_vcf_conversion.short.maf", sep = "\t", quote = FALSE, row.names = FALSE)
```

## Run maf2vcf.pl

#!/bin/bash

perl ~/Downloads/vcf2maf-1.6.17/maf2vcf.pl --input-maf /Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/input/snvs.full_cohort_minus_GZ229_and_GZ310.default_and_optimized.incl_silent_and_UTR.for_vcf_conversion.short.maf \
        --output-dir /Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/vcf/patient_level/excluding_GZ310_mutations \
        --ref-fasta ~/Downloads/vcf2maf-1.6.17/data/Homo_sapiens.GRCh37.75.dna.primary_assembly.fa \
        --per-tn-vcfs 1 \
        --tum-depth-col t_depth \
        --tum-rad-col t_ref_depth \
        --tum-vad-col t_var_depth \
        --nrm-depth-col n_depth \
        --nrm-rad-col n_ref_depth \
        --nrm-vad-col n_var_depth

## Analysis steps using the MutationalPatterns package

## NB: The de novo analysis is very computationally intensive and is not feasible to run on the local computer, so needs to be run on the projects space (numbers). Some of variables will have duplicate values, depending on whether there is an alternative on projects space.

## Load data

```{r}
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg19)
library(MutationalPatterns)

ref_genome <- "BSgenome.Hsapiens.UCSC.hg19"

# specify location of VCF files
vcf_path <- "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/vcf/patient_level/excluding_GZ310"

vcf_files <- list.files(path = vcf_path, pattern = "*.vcf", full.names = TRUE, all.files = TRUE)

# read in sample names and metadata to annotate the VCF files (remove GZ310 - sample id "GZ310T")
metadata <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/30_samples_final/metadata.txt", sep = "\t", header = TRUE)
# order sample names by the order in which the VCF files will be in (by patient_id column)
metadata <- metadata[order(metadata$tumor_id),]

# remove GZ310
metadata <- filter(metadata, metadata$tumor_id != "GZ-310")

sample_names <- metadata$tumor_id

# load VCF files
vcf <- read_vcfs_as_granges(vcf_files = vcf_files, sample_names = sample_names, ref_genome)
```

## Mutational characteristics

## Base substitution types

```{r}
library(gridExtra)

# retrieve base substitutions from the VCF GRanges object as REF>ALT
muts <- mutations_from_vcf(vcf[[1]])
# Convert base substitutions to the 6 types of conventional base types
types <- mut_type(vcf[[1]])
# retrieve the sequence context (one base upstream and one base downstream)
context <- mut_context(vcf[[1]], ref_genome)
# retrieve the type and context for all positions in the VCF GRanges object
type_context <- type_context(vcf[[1]], ref_genome)
# count mutation type occurrences for all VCF objects in the GRangesList
type_occurrences <- mut_type_occurrences(vcf, ref_genome)
```

## Mutation Spectrum

```{r}
# plot the mean relative contribution of each of the 6 base substitution types over all sampels
p1 <- plot_spectrum(type_occurrences)
# plot mutation spectrum with distinction between C>T at CpG sites and other sites
p2 <- plot_spectrum(type_occurrences, CT=TRUE)
# combine the plots
grid.arrange(p1, p2)

# facet by sample group - EBV
p4 <- plot_spectrum(type_occurrences, by = metadata$EBV_status, CT = TRUE, legend = TRUE)
# facet by sample group - path group
p5 <- plot_spectrum(type_occurrences, by = metadata$path_group, CT = TRUE, legend = TRUE)
# facet by sample group - mediastinal involvement
p6 <- plot_spectrum(type_occurrences, by = metadata$mediastinal_involvement, CT = TRUE, legend = TRUE)
grid.arrange(p4, p5, p6)
```

## 96 mutational profile

```{r}
# generate a 96-trinucleotide mutation count matrix
mut_mat <- mut_matrix(vcf_list = vcf, ref_genome = ref_genome)
# plot the 96 profile of all the samples
#plot_96_profile(mut_mat)
```

## Mutational signatures

Mutational signatures can be extracted from the mutation count matrix with NMF.  A critical parameter in NMF is the factorization rank, which is the number of mutational signatures.  You can find the optimal factorization rank using the NMF package. The most common approach is to choose the smallest rank for which the cophenetic correlation coefficient starts decreasing.

```{r}

## De novo mutational signature extraction using NMF

# first add a small pseudocount to mutation count matrix
mut_mat <- mut_mat + 0.0001
# use the NMF package to generate an estimate rank plot
estimate <- nmf(mut_mat, rank = 1:10, method = "brunet", nrun = 10, seed = 123456) # doesn't work on Rogue
# plot the estimates
plot(estimate)


# optimal rank appears to be 2
# perform a relative large number iterations to achieve stability and avoid local minima
# recommended number of iterations (runs) is 400-500 based on algorithm paper by Alexandrov et al.
nmf_res <- extract_signatures(mut_mat, rank = 2, nrun = 10)
# assign signature names
colnames(nmf_res$signatures) <- c("GZ Signature A", "GZ Signature B")
rownames(nmf_res$contribution) <- c("GZ Signature A", "GZ Signature B")
# print out contributions and signatures to output files
#write.table(nmf_res$contributions, "/data/projects/stacy/gzl_exomes/signature_analysis/output/28_samples_analysis/denovo-contributions.txt", sep = "\t", quote = FALSE)
#write.table(nmf_res$signature, "/data/projects/stacy/gzl_exomes/signature_analysis/output/28_samples_analysis/denovo-signatures.txt", sep = "\t", quote = FALSE)

# plot the 96-profile of the signatures
plot_96_profile(nmf_res$signatures)

# visualize the contribution of the signatures in a barplot
# order by specific groups of interest (contribution is a 2X28 matrix where 2 = signatures; 28 = samples)
col.order <- c("GZ-095", "GZ-BCC-020", "GZ-178", "GZ-048", "GZ-180", "GZ-BCC-013", "GZ-116", "GZ-099", "GZ-046", "GZ-BCC-054", "GZ-176", "GZ-184", "GZ-152", "GZ-044", "GZ-149", "GZ-092", "GZ-230", "GZ-086", "GZ-267", "GZ-068", "GZ-062", "GZ-197", "GZ-BCC-008", "GZ-235", "GZ-064", "GZ-294", "GZ-301", "GZ-032")
nmf_res$contribution <- nmf_res$contribution[, col.order]

# visualize the contribution in absolute number of mutations
plot_contribution(nmf_res$contribution, nmf_res$signature, mode = "absolute", coord_flip=TRUE)
# visualize the contribution in relative number of mutations
plot_contribution(nmf_res$contribution, nmf_res$signature, mode = "relative", coord_flip=TRUE)


# Plot signature contribution as a heatmap with sample clustering dendogram and specified signature order:
plot_contribution_heatmap(nmf_res$contribution, sig_order = c("GZ Signature A", "GZ Signature B"), cluster_samples=FALSE, plot_values=TRUE)
# calculate the relative contributions
denovo.contributions <- as.matrix(read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/28_samples_analysis/denovo-contributions.txt", sep = "\t", header = TRUE))
# transpose
denovo.contributions <- t(denovo.contributions)
# relative contribution
denovo.contrib.norm <- denovo.contributions / rowSums(denovo.contributions)
#write.table(denovo.contrib.norm, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/28_samples_analysis/denovo-contributions.norm.txt", sep = "\t", quote = FALSE)

# compare reconstructed mutational profile with original mutational profile
plot_compare_profiles(mut_mat[,1], nmf_res$reconstructed[,1], profile_names = c("Original", "Reconstructed"), condensed = TRUE)
```


## COSMIC mutational signatures

```{r}
# Download mutational signatures from the COSMIC website (saved local directory to avoid connection issues)
sp_url <- paste("https://cancer.sanger.ac.uk/cancergenome/assets/", "signatures_probabilities.txt", sep = "")
cancer_signatures = read.table(sp_url, sep = "\t", header = TRUE)

# Match the order of the mutation types to MutationalPatterns standard
new_order = match(row.names(mut_mat), cancer_signatures$Somatic.Mutation.Type)

# Reorder cancer signatures dataframe
cancer_signatures = cancer_signatures[as.vector(new_order),]

# Add trinucletiode changes names as row.names
row.names(cancer_signatures) = cancer_signatures$Somatic.Mutation.Type

# Keep only 96 contributions of the 30 signatures in matrix
cancer_signatures = as.matrix(cancer_signatures[,4:33])
```

## Similarity between de novo signatures and COSMIC signatures

```{r}
# calculate pairwise cosine similarity between mutational profiles and COSMIC signatures
cos_sim_denovo_signatures <- cos_sim_matrix(nmf_res$signatures, cancer_signatures)

# plot heatmap of cosine similarities
plot_cosine_heatmap(cos_sim_denovo_signatures, col_order = cosmic_order, cluster_rows = FALSE, plot_values=TRUE)

## plot signature contribution as heatmap
pheatmap1 <- plot_contribution_heatmap(nmf_res$contribution, cluster_samples=FALSE)
```

## Similarity between mutational profiles and COSMIC signatures

```{r}
# calculate pairwise cosine similarity between mutational profiles and COSMIC signatures
cos_sim_samples_signatures <- cos_sim_matrix(mut_mat, cancer_signatures)
# plot heatmap with specific signature order
plot_cosine_heatmap (cos_sim_samples_signatures, cluster_rows = FALSE)

# Find optimal contribution of COSMIC signatures to reconstruct 96 mutational profiles

# fit mutation matrix to the COSMIC mutational signatures
fit_res <- fit_to_signatures(mut_mat, cancer_signatures)
# print contributions to output
write.table(fit_res$contribution, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/contributions.txt", sep = "\t", quote = FALSE)
```

## Summary plots of top contributing signatures

```{r}
# plot optimal contribution of the COSMIC signatures in each sample as stacked barplot
# first select signatures with a minimum of X mutations (summed across all samples) contributing to that signature
select <- which (rowSums(fit_res$contribution) >= MIN_CONTRIBUTION)
# plot contribution barplot
plot_contribution(fit_res$contribution[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "absolute")

contribs <- as.data.frame.matrix(t(fit_res$contribution[select,]))
contribs$patient <- row.names(contribs)
contribs.long <- gather(contribs, signature, contribution, Signature.1:Signature.15)

# barplot (using absolute numbers of mutations)
p.absolute <- ggplot(data = contribs.long, 
             aes(x = patient, y = as.numeric(contribution), 
                     fill = signature)) + 
  geom_bar(stat = "identity", width = 0.7, size = 0.8) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_brewer(name = "Signature", palette="Set3") +
  xlab("") + 
  ylab("Number of mutations")


# and the same plot, but for relative contributions
plot_contribution(fit_res$contribution[select,], cancer_signatures[,select], coord_flip = TRUE, mode = "relative")

# and the customized version:

# Calculate relative proportions of mutations 

# first sum total contributions for each gene
contribs.long <- contribs.long %>% 
  dplyr::group_by(patient) %>% dplyr::mutate(total_contribution = sum(as.numeric(contribution)))
# calculate relative contribution per gene
contribs.long <- contribs.long %>% dplyr::mutate(relative_contribution = contribution / total_contribution)
# customized relative plot
p.relative <- ggplot(data = contribs.long, 
             aes(x = patient, y = as.numeric(relative_contribution), 
                     fill = signature)) + 
  geom_bar(stat = "identity", width = 0.7, size = 0.8) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_brewer(name = "Signature", palette="Set3") +
  xlab("") + 
  ylab("Number of mutations") +
  coord_flip()
```

## Create new contribution matrix based on grouping of key signatures

Categories: 
(1) Aging - #1
(2) Defective DNA mismatch repair (DDMR) - #3, 6, 15, 20, 26
(3) *Somatic hypermutation (SHM) - #2, 9, 13
(4) *POLE activity - #10
(5) Other COSMIC signatures with known aetiology (4, 7, 11, 22, 24, 29)
(6) COSMIC signatures with unknown aetiology

*New signatures added not used for the patient-level analysis
*AID/APOBEC and Polymerase-n activity were merged into the SHM group
**Previous signatures that have been omitted due to negligable contribution to gene mutations

```{r}
library(ggplot2)
library(tidyr)
library(plyr)
library(dplyr)

optimal_contributions <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/contributions.txt", sep = "\t", header = TRUE, row.names = 1, stringsAsFactors = FALSE)

# create new contribution matrix with just the 5 signature groups
df <- as.data.frame(t(optimal_contributions))

contributions.aging <- df$Signature.1
contributions.DDMR <- df$Signature.3 + df$Signature.6 + df$Signature.15 + df$Signature.20 + df$Signature.26
contributions.SHM <- df$Signature.2 + df$Signature.9 + df$Signature.13
contributions.POLE <- df$Signature.10
# all other signatures not captured above - other and "unknown"
contributions.other <- df$Signature.4 + df$Signature.7 + df$Signature.11 + df$Signature.22 + df$Signature.24 + df$Signature.29
contributions.unknown <- df$Signature.5 + df$Signature.8 + df$Signature.12 + df$Signature.14 + df$Signature.16 + df$Signature.17 + df$Signature.18 + df$Signature.19 + df$Signature.21 + df$Signature.23 + df$Signature.25 + df$Signature.27 + df$Signature.28 + df$Signature.30

contributions.grouped <- as.data.frame(cbind(row.names(df), contributions.aging, contributions.DDMR, contributions.SHM, contributions.POLE, contributions.other, contributions.unknown))

# convert from wide to long
contributions.grouped.long <- gather(contributions.grouped, 
                                     signature_group, 
                                     contribution, 
                                     contributions.aging:contributions.unknown)
colnames(contributions.grouped.long) <- c("patient", "signature_group", "contribution")

write.table(contributions.grouped.long, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/contributions-patient-level-normalized.grouped.txt", sep = "\t", quote = FALSE)

# Update ids so they contain dashes
contributions.grouped.long$patient <- gsub("\\.", "-", contributions.grouped.long$patient)
# Update signature group names
contributions.grouped.long$signature_group <- plyr::revalue(contributions.grouped.long$signature_group, 
                            c("contributions.aging"="Aging (1)", 
                            "contributions.DDMR"="Defective DNA mismatch repair (3, 6, 15, 20, 26)",
                            "contributions.SHM"="Somatic hypermutation (2, 9, 13)",
                            "contributions.POLE"="POLE activity (10)",
                            "contributions.other"="Other signatures with known aetiology (4, 7, 11, 22, 24, 29)",
                            "contributions.unknown"="Signatures with unknown aetiology"))

# add EBER status, mediastinal involvement, path group, and age group
# !!! ADD copy number !!!
colnames(contributions.grouped.long) <- c("tumor_id", "signature_group", "contribution")
contributions.grouped.long <- merge(contributions.grouped.long, metadata[c("tumor_id", "EBV_status", "mediastinal_involvement", "path_group", "age")], by = "tumor_id")
plot_data <- contributions.grouped.long

# map path groups to binary values (1/2 or 0/3)
plot_data <- plot_data %>% mutate(
     path_group_v2 = case_when(
         path_group == 1 | path_group == 2 ~ 1,
         path_group == 0 | path_group == 3 ~ 0
     )
)

# map mediastinal involvement to primary vs. non-primary (i.e. combine secondary, not involved)
plot_data <- plot_data %>% mutate(
     med_status = case_when(
         mediastinal_involvement == "Primary"  ~ 1,
         mediastinal_involvement == "Secondary" | mediastinal_involvement == "Not involved" ~ 0
     )
)

# Revalue age "< 45" to "<= 45"
plot_data$age <- plyr::revalue(plot_data$age, c("< 45"="<= 45"))

plot_data <- as.data.frame(plot_data)

########################
## EBV-specific plots ##
########################

# order patients by EBV status then by increasing numbers of mutations (need to group by patient)
# first sum total contributions for each patient
plot_data <- plot_data %>% 
  dplyr::group_by(tumor_id) %>% dplyr::mutate(total_contribution = sum(as.numeric(contribution)))
# calculate relative contribution per gene (useful for a relative contribution plot)
plot_data <- as.data.frame(plot_data %>% dplyr::mutate(relative_contribution = as.numeric(contribution) / total_contribution))
# now order first by EBV status, and then by decreasing total contribution by patient
plot_data <- plot_data[with(plot_data, order(as.character(EBV_status), -total_contribution)), ]

# specify order of patients (based on our sorting)
plot_data$tumor_id <- factor(plot_data$tumor_id)
tumors.ordered <- unique(plot_data$tumor_id)

# specify order of signature groups:
sig_groups.order <- c("Aging (1)",
                      "Defective DNA mismatch repair (3, 6, 15, 20, 26)",
                      "POLE activity (10)",
                      "Somatic hypermutation (2, 9, 13)",
                      "Other signatures with known aetiology (4, 7, 11, 22, 24, 29)",
                      "Signatures with unknown aetiology")

# barplot (using absolute numbers of mutations)
p.absolute <- ggplot(data = plot_data, 
             aes(factor(x = tumor_id, levels = tumors.ordered), 
                     y = as.numeric(contribution), 
                     fill = factor(signature_group, levels = sig_groups.order), 
                     color = EBV_status)) + 
  geom_bar(stat = "identity", width = 0.7, size = 0.8) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5)) +
  scale_colour_manual(name = "EBER status", 
                      values = c("Negative"="black", "Positive"="red")) +
  scale_fill_manual(name = "Signature group",
                    values = c("Aging (1)" = "rosybrown1",
                               "Defective DNA mismatch repair (3, 6, 15, 20, 26)" = "mediumpurple1",
                               "POLE activity (10)" = "lightgreen", 
                               "Somatic hypermutation (2, 9, 13)" = "dodgerblue", 
                               "Other signatures with known aetiology (4, 7, 11, 22, 24, 29)" = "gray",
                               "Signatures with unknown aetiology" = "gray48")) +
  xlab("") + 
  ylab("Number of mutations")
  

# now a slightly different plot, showing contributions by signature group (instead of patient), and encoded with EBER

# specify order of signature groups:
sig_groups.order <- c("Defective DNA mismatch repair (3, 6, 15, 20, 26)",
                      "Aging (1)",
                      "Signatures with unknown aetiology",
                      "Other signatures with known aetiology (4, 7, 11, 22, 24, 29)",
                      "Somatic hypermutation (2, 9, 13)",
                      "POLE activity (10)"
                      )

# barplot (using absolute numbers of mutations)
p.signatures <- ggplot(data = plot_data, 
             aes(factor(x = signature_group, levels = sig_groups.order), 
                     y = as.numeric(contribution), 
                     fill = EBV_status)) + 
  geom_bar(stat = "identity", width = 0.7, size = 0.8) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5)) +
  scale_fill_manual(name = "EBER status", 
                      values = c("Negative"="black", "Positive"="red")) +
  xlab("") + 
  ylab("Number of mutations") + 
  theme(legend.position="bottom") +
  scale_x_discrete(labels=c("Defective DNA mismatch repair (3, 6, 15, 20, 26)" = "DDMR", 
                            "Aging (1)" = "Aging",
                            "Signatures with unknown aetiology" = "Unknown",
                            "Other signatures with known aetiology (4, 7, 11, 22, 24, 29)" = "Other",
                            "Somatic hypermutation (2, 9, 13)" = "SHM",
                            "POLE activity (10)" = "POLE")) + 
  coord_flip()

# plot contributions by EBV status

plot_data.SHM <- filter(plot_data, plot_data$signature_group == "Somatic hypermutation (2, 9, 13)")

p.SHM <- ggplot(plot_data.SHM, aes(x=EBV_status, y=as.numeric(contribution) )) + 
  geom_boxplot() +
  ylab("Contribution of mutations to SHM signature") + 
  xlab("EBER status") +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.7)
  
#stat_summary(fun.y=mean, geom="point", shape=23, size=4)

ebv_pos <- filter(plot_data.SHM, EBV_status == "Positive")
ebv_neg <- filter(plot_data.SHM, EBV_status == "Negative")

# t-test to see if the distributions are significantly different
t.test(as.numeric(ebv_pos$contribution), as.numeric(ebv_neg$contribution))

#	Welch Two Sample t-test

#data:  as.numeric(ebv_pos$contribution) and as.numeric(ebv_neg$contribution)
#t = -1.9198, df = 24.934, p-value = 0.06639
#alternative hypothesis: true difference in means is not equal to 0
#95 percent confidence interval:
# -25.8010955   0.9076723
#sample estimates:
#mean of x mean of y 
# 5.971257 18.417969 
```

## Association tests (e.g. Are SHM signature mutations differentially distributed by pathology group?)

```{r}
MED_PRIMARY = 1
MED_NOT_PRIMARY = 0

PATH_GRP_1_2 = 1
PATH_GRP_0_3 = 0

AGE_45_UNDER = "<= 45"
AGE_45_TO_60 = "45 < age < 60"
AGE_60_PLUS = ">= 60"

signatures <- as.vector(unique(plot_data$signature_group))
patients <- as.vector(unique(plot_data$tumor_id))

NUM_signatures <- length(signatures)
NUM_patients <- length(patients)

contingency_table <- array(dim = c(NUM_signatures*2, NUM_patients*2))
i <- 1
j <- 1

pval_table <- array(dim = c(NUM_signatures, NUM_patients))
k <- 1
l <- 1

# for each gene
for (patient_id in patients) {
  
  # for each signature
  for (signature_name in signatures){
    
    # get values for contingency table
    num.patient_and_sig <- sum(as.numeric((filter(plot_data, 
                                    tumor_id == patient_id & signature_group == signature_name))$contribution))
    num.not_patient_and_sig <- sum(as.numeric((filter(plot_data, 
                                    tumor_id != patient_id & signature_group == signature_name))$contribution)) 
    num.patient_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    tumor_id == patient_id & signature_group != signature_name))$contribution)) 
    num.not_patient_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    tumor_id != patient_id & signature_group != signature_name))$contribution)) 
    
    # chi-square test
    sam <- array(dim = c(2,2))
    sam[1, ] <- c(num.patient_and_sig, num.not_patient_and_sig)
    sam[2, ] <- c(num.patient_and_not_sig, num.not_patient_and_not_sig)
    
    curr_row.start <- i
    curr_row.end <- i+1
    curr_col.start <- j
    curr_col.end <- j+1
    
    contingency_table[curr_row.start:curr_row.end, curr_col.start:curr_col.end] <- sam
    i <- i + 2
    
    # Chi-Square test
    pvalue <- chisq.test(sam)$p.value
    pval_table[k, l] <- pvalue
    k <- k + 1
  }
  # continue iterating through genes
  j <- j + 2
  l <- l + 1
  # reset signatures
  i <- 1
  k <- 1
}
  
colnames(pval_table) <- patients
row.names(pval_table) <- signatures

not.patients <- paste("Not ", patients, sep = "")
contingency.col_names <- c(rbind(patients, not.patients))
not.signatures <- paste("Not ", signatures, sep = "")
contingency.row_names <- c(rbind(signatures, not.signatures))

colnames(contingency_table) <- contingency.col_names
row.names(contingency_table) <- contingency.row_names

#write.table(contingency_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/patient_mutations_vs_key_cosmic_signatures-contingency_table.txt", sep = "\t", quote = FALSE)

#write.table(pval_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/chisq-patient_mutations_vs_key_cosmic_signatures.txt", sep = "\t", quote = FALSE)

contingency_table.patient_groups <- contingency_table
pvalue_table.patient_groups <- pval_table
```

## Associations of signatures with EBV-status

```{r}
signatures <- as.vector(unique(plot_data$signature_group))
EBV_stati <- as.vector(unique(plot_data$EBV_status))

NUM_signatures <- length(signatures)
NUM_EBV_stati <- length(EBV_stati)

contingency_table <- array(dim = c(NUM_signatures*2, NUM_EBV_stati*2))
i <- 1
j <- 1

pval_table <- array(dim = c(NUM_signatures, NUM_EBV_stati))
k <- 1
l <- 1

# for each gene
for (EBV_stat in EBV_stati) {
  
  # for each signature
  for (signature_name in signatures){
    
    # get values for contingency table
    num.EBV_stat_and_sig <- sum(as.numeric((filter(plot_data, 
                                    EBV_status == EBV_stat & signature_group == signature_name))$contribution))
    num.not_EBV_stat_and_sig <- sum(as.numeric((filter(plot_data, 
                                    EBV_status != EBV_stat & signature_group == signature_name))$contribution)) 
    num.EBV_stat_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    EBV_status == EBV_stat & signature_group != signature_name))$contribution)) 
    num.not_EBV_stat_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    EBV_status != EBV_stat & signature_group != signature_name))$contribution)) 
    
    # chi-square test
    sam <- array(dim = c(2,2))
    sam[1, ] <- c(num.EBV_stat_and_sig, num.not_EBV_stat_and_sig)
    sam[2, ] <- c(num.EBV_stat_and_not_sig, num.not_EBV_stat_and_not_sig)
    
    curr_row.start <- i
    curr_row.end <- i+1
    curr_col.start <- j
    curr_col.end <- j+1
    
    contingency_table[curr_row.start:curr_row.end, curr_col.start:curr_col.end] <- sam
    i <- i + 2
    
    # Chi-Square test
    pvalue <- chisq.test(sam)$p.value
    pval_table[k, l] <- pvalue
    k <- k + 1
  }
  # continue iterating through genes
  j <- j + 2
  l <- l + 1
  # reset signatures
  i <- 1
  k <- 1
}
  
colnames(pval_table) <- EBV_stati
row.names(pval_table) <- signatures

not.EBV_stati <- paste("Not ", EBV_stati, sep = "")
contingency.col_names <- c(rbind(EBV_stati, not.EBV_stati))
not.signatures <- paste("Not ", signatures, sep = "")
contingency.row_names <- c(rbind(signatures, not.signatures))

colnames(contingency_table) <- contingency.col_names
row.names(contingency_table) <- contingency.row_names

#write.table(contingency_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/EBV_status_vs_key_cosmic_signatures-contingency_table.txt", sep = "\t", quote = FALSE)

#write.table(pval_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/chisq-EBV_status_vs_key_cosmic_signatures.txt", sep = "\t", quote = FALSE)

contingency_table.ebv_groups <- contingency_table
pvalue_table.ebv_groups <- pval_table
```

## Associations of signatures with path group

```{r}
signatures <- as.vector(unique(plot_data$signature_group))
path_groups <- as.vector(unique(plot_data$path_group_v2))

NUM_signatures <- length(signatures)
NUM_path_groups <- length(path_groups)

contingency_table <- array(dim = c(NUM_signatures*2, NUM_path_groups*2))
i <- 1
j <- 1

pval_table <- array(dim = c(NUM_signatures, NUM_path_groups))
k <- 1
l <- 1

# for each gene
for (path_group in path_groups) {
  
  # for each signature
  for (signature_name in signatures){
    
    # get values for contingency table
    num.path_group_and_sig <- sum(as.numeric((filter(plot_data, 
                                    path_group_v2 == path_group & signature_group == signature_name))$contribution))
    num.not_path_group_and_sig <- sum(as.numeric((filter(plot_data, 
                                    path_group_v2 != path_group & signature_group == signature_name))$contribution)) 
    num.path_group_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    path_group_v2 == path_group & signature_group != signature_name))$contribution)) 
    num.not_path_group_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    path_group_v2 != path_group & signature_group != signature_name))$contribution)) 
    
    # chi-square test
    sam <- array(dim = c(2,2))
    sam[1, ] <- c(num.path_group_and_sig, num.not_path_group_and_sig)
    sam[2, ] <- c(num.path_group_and_not_sig, num.not_path_group_and_not_sig)
    
    curr_row.start <- i
    curr_row.end <- i+1
    curr_col.start <- j
    curr_col.end <- j+1
    
    contingency_table[curr_row.start:curr_row.end, curr_col.start:curr_col.end] <- sam
    i <- i + 2
    
    # Chi-Square test
    pvalue <- chisq.test(sam)$p.value
    pval_table[k, l] <- pvalue
    k <- k + 1
  }
  # continue iterating through genes
  j <- j + 2
  l <- l + 1
  # reset signatures
  i <- 1
  k <- 1
}
  
colnames(pval_table) <- path_groups
row.names(pval_table) <- signatures

not.path_groups <- paste("Not ", path_groups, sep = "")
contingency.col_names <- c(rbind(path_groups, not.path_groups))
not.signatures <- paste("Not ", signatures, sep = "")
contingency.row_names <- c(rbind(signatures, not.signatures))

colnames(contingency_table) <- contingency.col_names
row.names(contingency_table) <- contingency.row_names

#write.table(contingency_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/path_groups_vs_key_cosmic_signatures-contingency_table.txt", sep = "\t", quote = FALSE)

#write.table(pval_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/chisq-path_groups_vs_key_cosmic_signatures.txt", sep = "\t", quote = FALSE)

contingency_table.path_groups <- contingency_table
pvalue_table.path_groups <- pval_table
```

## Associations of signatures with mediastinal involvement

```{r}
signatures <- as.vector(unique(plot_data$signature_group))
med_groups <- as.vector(unique(plot_data$med_status))

NUM_signatures <- length(signatures)
NUM_med_groups <- length(med_groups)

contingency_table <- array(dim = c(NUM_signatures*2, NUM_med_groups))
i <- 1

pval_table <- array(dim = c(NUM_signatures, 1))
k <- 1

# for each signature
for (signature_name in signatures){
    
    # get values for contingency table
    num.med_primary_and_sig <- sum(as.numeric((filter(plot_data, 
                                    med_status == 1 & signature_group == signature_name))$contribution))
    num.med_primary_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    med_status == 1 & signature_group != signature_name))$contribution)) 
    num.med_not_primary_and_sig <- sum(as.numeric((filter(plot_data, 
                                    med_status == 0 & signature_group == signature_name))$contribution)) 
    num.med_not_primary_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    med_status == 0 & signature_group != signature_name))$contribution)) 
    
    
    # chi-square test
    sam <- array(dim = c(2,2))
    sam[1, ] <- c(num.med_primary_and_sig, num.med_not_primary_and_sig)
    sam[2, ] <- c(num.med_primary_and_not_sig, num.med_not_primary_and_not_sig)

    curr_row.start <- i
    curr_row.end <- i+1

    contingency_table[curr_row.start:curr_row.end, 1:2] <- sam
    i <- i + 2
    
    # Chi-Square test
    pvalue <- chisq.test(sam)$p.value
    pval_table[k, 1] <- pvalue
    k <- k + 1
}

colnames(pval_table) <- c("mediastinal_involvement")
row.names(pval_table) <- signatures

not.signatures <- paste("Not ", signatures, sep = "")
contingency.row_names <- c(rbind(signatures, not.signatures))

colnames(contingency_table) <- med_groups
row.names(contingency_table) <- contingency.row_names


write.table(contingency_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/med_groups_vs_key_cosmic_signatures-contingency_table.txt", sep = "\t", quote = FALSE)

write.table(pval_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/chisq-med_groups_vs_key_cosmic_signatures.txt", sep = "\t", quote = FALSE)

contingency_table.med_groups <- contingency_table
pvalue_table.med_groups <- pval_table
```


## Associations of signatures with age groups

```{r}
signatures <- as.vector(unique(plot_data$signature_group))
age_groups <- as.vector(unique(plot_data$age))

NUM_signatures <- length(signatures)
NUM_age_groups <- length(age_groups)

contingency_table <- array(dim = c(NUM_signatures*2, NUM_age_groups))
i <- 1

pval_table <- array(dim = c(NUM_signatures, 1))
k <- 1

# for each signature
for (signature_name in signatures){
    
    # get values for contingency table
    num.under_45_and_sig <- sum(as.numeric((filter(plot_data, 
                                    age == "<= 45" & signature_group == signature_name))$contribution))
    num.under_45_and_no_sig <- sum(as.numeric((filter(plot_data, 
                                    age == "<= 45" & signature_group != signature_name))$contribution)) 
    num.45_to_60_and_sig <- sum(as.numeric((filter(plot_data, 
                                    age == "45 < age < 60" & signature_group == signature_name))$contribution)) 
    num.45_to_60_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    age == "45 < age < 60" & signature_group != signature_name))$contribution)) 
    num.over_60_and_sig <- sum(as.numeric((filter(plot_data, 
                                    age == ">= 60" & signature_group == signature_name))$contribution))
    num.over_60_and_not_sig <- sum(as.numeric((filter(plot_data, 
                                    age == ">= 60" & signature_group != signature_name))$contribution)) 
    
    # chi-square test
    sam <- array(dim = c(2,3))
    sam[1, ] <- c(num.under_45_and_sig, num.45_to_60_and_sig, num.over_60_and_sig)
    sam[2, ] <- c(num.under_45_and_no_sig, num.45_to_60_and_not_sig, num.over_60_and_not_sig)
    
    curr_row.start <- i
    curr_row.end <- i+1

    contingency_table[curr_row.start:curr_row.end, 1:3] <- sam
    i <- i + 2
    
    # Chi-Square test
    pvalue <- chisq.test(sam)$p.value
    pval_table[k, 1] <- pvalue
    k <- k + 1
}

colnames(pval_table) <- c("age_group")
row.names(pval_table) <- signatures

not.signatures <- paste("Not ", signatures, sep = "")
contingency.row_names <- c(rbind(signatures, not.signatures))

colnames(contingency_table) <- age_groups
row.names(contingency_table) <- contingency.row_names

write.table(contingency_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/age_groups_vs_key_cosmic_signatures-contingency_table.txt", sep = "\t", quote = FALSE)

write.table(pval_table, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/chisq-age_groups_vs_key_cosmic_signatures.txt", sep = "\t", quote = FALSE)

contingency_table.age_groups <- contingency_table
pvalue_table.age_groups <- pval_table
```

## Combine contingency tables and p-value tables

```{r}
contingency_table.combined <- cbind(contingency_table.ebv_groups,
                                    contingency_table.med_groups,
                                    contingency_table.path_groups,
                                    contingency_table.age_groups)

pvalue_table.combined <- cbind(pvalue_table.ebv_groups,
                               pvalue_table.med_groups,
                               pvalue_table.path_groups,
                               pvalue_table.age_groups)

write.table(contingency_table.combined, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/ALL_groups_vs_key_cosmic_signatures-contingency_table.txt", sep = "\t", quote = FALSE)

write.table(pvalue_table.combined, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/patient_level-excl_GZ310/chisq_tests/chisq-ALL_groups_vs_key_cosmic_signatures.txt", sep = "\t", quote = FALSE)

```

## Calculate enrichment of the AID/APOBEC signature in EBV vs. EBV- groups

```{r}
library(dplyr)

EBV_pos.AID_sig.data <- filter(plot_data, EBV_status == "Positive" & signature_group == "Somatic hypermutation (2, 9, 13)")
EBV_pos.not_AID_sig.data <- filter(plot_data, EBV_status == "Positive" & !(signature_group == "Somatic hypermutation (2, 9, 13)"))
EBV_neg.AID_sig.data <- filter(plot_data, EBV_status == "Negative" & signature_group == "Somatic hypermutation (2, 9, 13)")
EBV_neg.not_AID_sig.data <- filter(plot_data, EBV_status == "Negative" & !(signature_group == "Somatic hypermutation (2, 9, 13)"))

EBV_pos.AID_sig.num <- sum(as.numeric(EBV_pos.AID_sig.data$contribution)) # 42
EBV_pos.not_AID_sig.num <- sum(as.numeric(EBV_pos.not_AID_sig.data$contribution)) # 608
EBV_neg.AID_sig.num <- sum(as.numeric(EBV_neg.AID_sig.data$contribution)) # 405
EBV_neg.not_AID_sig.num <- sum(as.numeric(EBV_neg.not_AID_sig.data$contribution)) # 3392

# chi-square test
sam <- array(dim = c(2,2))
sam[1, ] <- c(EBV_pos.AID_sig.num, EBV_neg.AID_sig.num)
sam[2, ] <- c(EBV_pos.not_AID_sig.num, EBV_neg.not_AID_sig.num)
  
# Chi-Square test to assess if AID/APOBEC in EBV+ cases is more different than expected
pvalue <- chisq.test(sam)$p.value # 0.001134251

```

