---
title: "GZL_exomes-signature_analysis-chisq"
author: "Stacy Hung"
date: "September 10, 2019"
output: html_document
---

There are two general comparisons that we want to make:

1. "Absolute" number of mutations associated with a given signature vs. number of mutations occurring in target panel gene (e.g. SOCS1)
2. "Absolute" number of patients associated with a given signature vs. number of patients with a particular clinical phenotype (e.g. EBV+ / EBV-)

***Here "absolute" is a very tenuous term used here since this number is based on the highest probability (e.g. each patient's mutational signature is attributed to potentially all 30 COSMIC signatures, with each signature having a different probability - then, the signature with the highest probability is assigned as the "associated" signature for the patient; thus, in this context, all other signatures are ignored).  The assumption here is that the top scoring probability values are very high (high enough that there is substantial doubt the patient would be associated with any other signature).  As this assumption is very unlikely to be biologically true, we need to be very careful and specific with the wording of the results - e.g. "Patients with a majority of mutations likely attributed to the AID/APOBEC signature were significantly linked to EBV+ patients".



## To modify for contigency table containing mutation frequency data for each gene, and associated subgroups (e.g mediastinal status, EBV status, path group, etc.)

```{r}

contributions <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/mutationalPatterns/output/30_samples/contributions-for_contingency.txt", sep = "\t", header = TRUE)

# first map COSMIC signatures to descriptive names for signatures of interest

# 1. Defective DNA mismatch repair (Signatures #3, 6, 15, 20, 26)
contributions$signature_DDMR <- contributions$Signature.3 + contributions$Signature.6 + contributions$Signature.15 + 
  contributions$Signature.20 + contributions$Signature.26

# 2. AID / APOBEC activity (Signatures 2, 13)
contributions$signature_AID <- contributions$Signature.2 + contributions$Signature.13

# 3. Polymerase-n-mediated somatic hypermutation (Signature #9)
contributions$signature_hyper <- contributions$Signature.9

# 4. Aging (Signature 1)
contributions$signature_aging <- contributions$Signature.1

# 5. All other cosmic signatures
cosmic_signatures <- colnames(contributions[,11:40])
contributions$signature_other <- rowSums(contributions[, cosmic_signatures]) - 
  contributions$signature_DDMR - contributions$signature_AID - contributions$signature_hyper - contributions$signature_aging

# extract relevant columns
keeps <- c(colnames(contributions[,1:10]), colnames(contributions[,41:45]))

contributions.subset <- contributions[, keeps]

# convert from wide to long (group signatures separately from patient characteristics)
library(tidyr)
contributions.long <- gather(contributions.subset, signature, contribution, signature_DDMR:signature_other, factor_key = TRUE)

write.table(contributions.subset, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/30_samples_final/contributions_with_clinical_features.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

## Try applying the chi-square test to a very specific example involving patient-level COSMIC contribution and EBV status

```{r}
library(dplyr)

top_contrib <- contributions.long %>% group_by(case_id) %>% top_n(1, contribution)
write.table(top_contrib, "/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/30_samples_final/top_contributions_with_clinical_features.txt", sep = "\t", row.names = FALSE, quote = FALSE)

# contingency table for signature vs. EBV status
EBV_table <- table(top_contrib$signature, top_contrib$EBER_pos)
EBV_table.df <- as.data.frame.matrix(EBV_table)
colnames(EBV_table.df) <- c("EBV_neg", "EBV_pos")
# remove rows with no data:
EBV_table.df.no0 = EBV_table.df[rowSums(EBV_table.df)!=0, ]
# apply chi-square test to contingency table
EBV.chisq <- chisq.test(EBV_table.df.no0)
# assess specific signatures
# EBV vs. DDMR signature
EBV_table.vs.DDMR <- as.data.frame.array(t(EBV_table.df.no0))
EBV_table.vs.DDMR$signature_not_DDMR <- EBV_table.vs.DDMR$signature_aging + EBV_table.vs.DDMR$signature_other
EBV_table.vs.DDMR$signature_aging <- NULL
EBV_table.vs.DDMR$signature_other <- NULL
EBV_vs_DDMR.chisq <- chisq.test(EBV_table.vs.DDMR)
# EBV vs. aging signature
EBV_vs_aging.table <- as.data.frame.array(t(EBV_table.df.no0))
EBV_vs_aging.table$signature_not_aging <- EBV_vs_aging.table$signature_DDMR + EBV_vs_aging.table$signature_other
EBV_vs_aging.table$signature_DDMR <- NULL
EBV_vs_aging.table$signature_other <- NULL
EBV_vs_aging.chisq <- chisq.test(EBV_vs_aging.table)
# EBV vs. other signatures
EBV_vs_other.table <- as.data.frame.array(t(EBV_table.df.no0))
EBV_vs_other.table$signature_not_other <- EBV_vs_other.table$signature_DDMR + EBV_vs_other.table$signature_aging
EBV_vs_other.table$signature_DDMR <- NULL
EBV_vs_other.table$signature_aging <- NULL
EBV_vs_other.chisq <- chisq.test(EBV_vs_other.table)


# contingency table for signature vs. mediastinal status
med_table <- table(top_contrib$signature, top_contrib$Mediastinal_primary)
med_table.df <- as.data.frame.matrix(med_table)
colnames(med_table.df) <- c("mediastinal_nonprimary", "mediastinal_primary")
# remove rows with no data
med_table.df.no0 = med_table.df[rowSums(med_table.df)!=0, ]
# apply chi-square test to contingency table
med_status.chisq <- chisq.test(med_table.df.no0)
# assess specific signatures
# mediastinal status vs. DDMR signature
med_status_vs_DDMR.table <- as.data.frame(t(med_table.df.no0))
med_status_vs_DDMR.table$signature_not_DDMR <- med_status_vs_DDMR.table$signature_aging + med_status_vs_DDMR.table$signature_other
med_status_vs_DDMR.table$signature_aging <- NULL
med_status_vs_DDMR.table$signature_other <- NULL
med_status_vs_DDMR.chisq <- chisq.test(med_status_vs_DDMR.table)
# mediastinal status vs. aging signature
med_status_vs_aging.table <- as.data.frame(t(med_table.df.no0))
med_status_vs_aging.table$signature_not_aging <- med_status_vs_aging.table$signature_DDMR + med_status_vs_aging.table$signature_other
med_status_vs_aging.table$signature_DDMR <- NULL
med_status_vs_aging.table$signature_other <- NULL
med_status_vs_aging.chisq <- chisq.test(med_status_vs_aging.table)
# mediastinal status vs. other signatures
med_status_vs_other.table <- as.data.frame(t(med_table.df.no0))
med_status_vs_other.table$signature_not_other <- med_status_vs_other.table$signature_DDMR + med_status_vs_other.table$signature_aging
med_status_vs_other.table$signature_DDMR <- NULL
med_status_vs_other.table$signature_aging <- NULL
med_status_vs_other.chisq <- chisq.test(med_status_vs_other.table)


# contingency table for signature vs. path group
path_table <- table(top_contrib$signature, top_contrib$Path_group_1_2)
path_table.df <- as.data.frame.matrix(path_table)
colnames(path_table.df) <- c("Path_group_0_3", "Path_group_1_2")
# remove rows with no data
path_table.df.no0 <- path_table.df[rowSums(path_table.df)!=0, ]
# apply chi-square test to contingency table
path_grp.chisq <- chisq.test(path_table.df.no0)
# assess specific signatures
#  path group vs. DDMR signature
path_grp_vs_DDMR.table <- as.data.frame(t(path_table.df.no0))
path_grp_vs_DDMR.table$signature_not_DDMR <- path_grp_vs_DDMR.table$signature_aging + path_grp_vs_DDMR.table$signature_other
path_grp_vs_DDMR.table$signature_aging <- NULL
path_grp_vs_DDMR.table$signature_other <- NULL
path_grp_vs_DDMR.chisq <- chisq.test(path_grp_vs_DDMR.table)
#  path group vs. aging signature
path_grp_vs_aging.table <- as.data.frame(t(path_table.df.no0))
path_grp_vs_aging.table$signature_not_aging <- path_grp_vs_aging.table$signature_DDMR + path_grp_vs_aging.table$signature_other
path_grp_vs_aging.table$signature_DDMR <- NULL
path_grp_vs_aging.table$signature_other <- NULL
path_grp_vs_aging.chisq <- chisq.test(path_grp_vs_aging.table)
#  path group vs. other signatures
path_grp_vs_other.table <- as.data.frame(t(path_table.df.no0))
path_grp_vs_other.table$signature_not_other <- path_grp_vs_other.table$signature_DDMR + path_grp_vs_other.table$signature_aging
path_grp_vs_other.table$signature_aging <- NULL
path_grp_vs_other.table$signature_DDMR <- NULL
path_grp_vs_other.chisq <- chisq.test(path_grp_vs_other.table)


# contingency table for signature vs. Age group
age_grp_table <- read.table("/Volumes/shung/projects/gzl_exomes/snv_analysis-20181127/signature_analysis/analysis/30_samples_final/cosmic_vs_age_grp.txt", sep = "\t", header = TRUE, fill = TRUE, row.names = 1)
# apply chi-square test to contingency table
age_grp.chisq <- chisq.test(age_grp_table)
# assess specific signatures
# age group vs. DDMR signature
age_grp_vs_DDMR.table <- as.data.frame(t(age_grp_table))
age_grp_vs_DDMR.table$signature_not_DDMR <- age_grp_vs_DDMR.table$signature_aging + age_grp_vs_DDMR.table$signature_other
age_grp_vs_DDMR.table$signature_aging <- NULL
age_grp_vs_DDMR.table$signature_other <- NULL
age_grp_vs_DDMR.chisq <- chisq.test(age_grp_vs_DDMR.table)
# age group vs. aging signature
age_grp_vs_aging.table <- as.data.frame(t(age_grp_table))
age_grp_vs_aging.table$signature_not_aging <- age_grp_vs_aging.table$signature_DDMR + age_grp_vs_aging.table$signature_other
age_grp_vs_aging.table$signature_DDMR <- NULL
age_grp_vs_aging.table$signature_other <- NULL
age_grp_vs_aging.chisq <- chisq.test(age_grp_vs_aging.table)
# age group vs. other signatures
age_grp_vs_other.table <- as.data.frame(t(age_grp_table))
age_grp_vs_other.table$signature_not_other <- age_grp_vs_other.table$signature_DDMR + age_grp_vs_other.table$signature_aging
age_grp_vs_other.table$signature_DDMR <- NULL
age_grp_vs_other.table$signature_aging <- NULL
age_grp_vs_other.chisq <- chisq.test(age_grp_vs_other.table)
```

